---
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r}
here_path <- here::here()
library(extrafont)
loadfonts()
library(ggplot2)
theme_set(theme_minimal(base_family = "PT Sans"))
```


R tools to access open data from Eurostat database
=========================================

# Search and download

Data in the Eurostat database is stored in tables.
Each table has an identifier, a short table_code, and 
a description (e.g. tps00199 - Total fertility rate).

Key eurostat functions allow to find the table_code, 
download the eurostat table and polish labels in the table.

## Find the table code

The search_eurostat(pattern, ...) function scans the directory 
of Eurostat tables and returns codes and descriptions of 
tables that match pattern.

```{r}
library("eurostat")
query <- search_eurostat(pattern = "fertility rate", 
                         type = "table", fixed = FALSE)
query[,1:2]
## title                                 code    
## <chr>                                 <chr>   
## Total fertility rate by NUTS 2 region tgs00100
## Total fertility rate                  tps00199
## Total fertility rate by NUTS 2 region tgs00100
```


## Download the table

The get_eurostat(id, time_format = "date", filters = "none", 
type = "code", cache = TRUE, ...) function downloads the 
requested table from the Eurostat bulk download facility or
from The Eurostat Web Services JSON API (if filters are defined). 
Downloaded data is cached (if cache=TRUE). 
Additional arguments define how to read the time column
(time_format) and if table dimensions shall be kept as 
codes or converted to labels (type).

```{r}
ct <- c("AT","BE","BG","CH","CY","CZ","DE","DK","EE","EL","ES",
        "FI","FR","HR","HU","IE","IS","IT","LI","LT","LU","LV",
        "MT","NL","NO","PL","PT","RO","SE","SI","SK","UK")
dat <- get_eurostat(id="tps00199", time_format="num", 
                    filters = list(geo = ct))
dat[1:2,]
## indic_de geo    time values
## TOTFERRT AT     2006   1.41
## TOTFERRT AT     2007   1.38
```

## Add labels

The label_eurostat(x, lang = "en", ...) gets definitions for Eurostat
codes and replace them with labels in given language ("en", "fr" or "de")

```{r}
dat <- label_eurostat(dat)
dat[1:3,]
## indic_de             geo      time values
## <fct>                <fct>   <dbl>  <dbl>
## Total fertility rate Austria  2006   1.41
## Total fertility rate Austria  2007   1.38
## Total fertility rate Austria  2008   1.42
```


## eurostat and plots

The get_eurostat() function returns tibbles in the long format. Packages
dplyr and tidyr are well suited to transform these objects. The ggplot2
package is well suited to plot these objects. 

```{r}
dat <- get_eurostat(id="tps00199", filters = list(geo = cntries))
library(ggplot2)
library(dplyr)
ggplot(dat, 
       aes(x = time, y = values, color = geo, label = geo)) +
  geom_line(alpha = .5) +
  geom_text(data = dat %>% group_by(geo) %>% 
              filter(time == max(time)), 
            size = 2.6) + 
  theme(legend.position = "none") +
  labs(title = "Total fertility rate, 2006-2017", 
       x = "Year", y = "%") -> p
# save plot
ggsave(filename = glue::glue("{here_path}/inst/extras/cheatsheet/lineplot.pdf"),plot = p, width = 4.84, height = 3.26, device = cairo_pdf)
```

```{r}
dat_2015 <- dat %>% 
  filter(time == "2015-01-01") 
ggplot(dat_2015, aes(x = reorder(geo, values), y = values)) +
  geom_col(color = "white", fill = "grey80") + 
  theme(axis.text.x = element_text(size = 6)) +
  labs(title = "Total fertility rate, 2015",
       y = "%", x = NULL) -> p
# save plot
ggsave(filename = glue::glue("{here_path}/inst/extras/cheatsheet/barplot.pdf"),plot = p, width = 4.84, height = 3.26, device = cairo_pdf)
```


# eurostat and maps

## Fetch and process data

There are three function to work with geospatial data from GISCO. The
get_eurostat_geospatial() returns spatial data as sf-object. 
Object can me merged with data.frames using dplyr::*_join()-functions.
The cut_to_classes() is a wrapper for cut() - function and 
is used for categorizing data for maps with tidy labels.

```{r}
mapdata <- get_eurostat_geospatial(nuts_level = 0) %>% 
  left_join(dat_2015) %>% 
  mutate(cat = cut_to_classes(values, n=4, decimals=1))
head(select(mapdata,geo,values,cat))
## geo values        cat                       geometry
## BG   1.53 1.5 ~< 1.7 MULTIPOLYGON (((22.99717 43...
## CH   1.54 1.5 ~< 1.7 MULTIPOLYGON (((8.61383 47....
## CY   1.32 1.3 ~< 1.5 MULTIPOLYGON (((33.75237 34...
## AL   1.59 1.5 ~< 1.7 MULTIPOLYGON (((19.831 42.4...
## CZ   1.57 1.5 ~< 1.7 MULTIPOLYGON (((14.49122 51...
## BE   1.70 1.5 ~< 1.7 MULTIPOLYGON (((5.10218 51....
```


## Plot a map

The sf-object returned are ready to be plotted with 
ggplot::geom_sf()-function.

```{r}
ggplot(mapdata, aes(fill = cat)) +
  scale_fill_brewer(palette = "RdYlBu") +
  geom_sf(color = alpha("white",1/3), alpha = .6) + 
  xlim(c(-12,44)) + ylim(c(35,70)) +
  labs(title = "Total fertility rate, 2015",
       subtitle = "Avg. number of life births per woman", 
       fill = "%") -> p
# save plot
ggsave(filename = glue::glue("{here_path}/inst/extras/cheatsheet/mapplot.pdf"),plot = p, width = 5.84, height = 4.26, device = cairo_pdf)
```

*******

This onepager presents the eurostat package Leo Lahti, Janne Huovari, 
Markus Kainu, Przemyslaw Biecek 2014-2019 
package version 3.3.55 URL: https://github.com/rOpenGov/eurostat

See also: Lahti et al. Retrieval and Analysis of Eurostat Open Data
with the eurostat Package. The R Journal 2017.

CC BY Przemyslaw Biecek & Markus Kainu
https://creativecommons.org/licenses/by/4.0/
