<!--
%\VignetteEngine{knitr}
%\VignetteIndexEntry{An R Markdown Vignette made with knitr}
-->

Eurostat R tools
===========

This R package provides tools to access open data from [Eurostat](http://epp.eurostat.ec.europa.eu/portal/page/portal/statistics/themes). 

The present implementation retrieves the Eurostat data through the [Statistics
Finland](http://www.stat.fi/tup/tilastotietokannat/index_fi.html) data portal. Pull requests for extensions are welcome.

This R package is part of the [rOpenGov](http://ropengov.github.io)
project.

The listings of [Eurostat](http://www.stat.fi/org/lainsaadanto/avoin_data.html) data sets are available for browsing in [PCAxis](http://pxweb2.stat.fi/Database/Eurostat/databasetree_fi.asp) and [CSV](http://pxweb2.stat.fi/database/Eurostatn/Eurostatn_rap.csv) formats.


## Installation

Release version for general use:

```{r install, eval=FALSE}
install.packages("eurostat")
library(eurostat)
```

Development version (potentially unstable):

```{r install2, eval=FALSE}
install.packages("devtools")
library(devtools)
install_github("eurostat", "ropengov")
library(eurostat)
```


### Browsing the data

List of the data sets is available in R:

```{r tilastokeskus-1, warning=FALSE, message=FALSE}
# Load the library
library(eurostat)

# Eurostat open data listing
datasets.eurostat <- list_eurostat_files()

# Descriptions of the first entries
head(datasets.eurostat$DESCRIPTION)

# Investigate the first entry 
print(datasets.eurostat[1,])
```

## Retrieving the data

Retrieve data from Eurostat by defining URL of the data set. For the
listing of available data sets and their corresponding URLs, see
above.

```{r tilastokeskus-2, warning=FALSE, message=FALSE}
library(eurostat)

# Define URL (see list_eurostat_files())
url <- "http://pxweb2.stat.fi/Database/Eurostat/ymp/t2020_30.px"

# Download the data
df <- get_eurostat(url)

# Inspect the first entries..
df[1:3,]
```


## Related tools

The [datamart](http://cran.r-project.org/web/packages/datamart/index.html) package contains related tools to access Eurostat data sets. At the time of writing this tutorial this package seems, however, to be in an experimental stage.

The SmarterPoland package has good tools for Eurostat:

```{r smarterpoland, warning=FALSE, message=FALSE}
library(SmarterPoland)
# info about passagers
grepEurostatTOC("split of passenger transport")
## get table
tmp <- getEurostatRCV("tsdtr210")
summary(tmp)
```


## Licensing and Citations

### Citing the Data

Regarding the data, kindly cite [Eurostat](http://epp.eurostat.ec.europa.eu/portal/page/portal/statistics/search_database). 


### Citing the R tools

This work can be freely used, modified and distributed under the
[Two-clause FreeBSD
license](http://en.wikipedia.org/wiki/BSD\_licenses). Kindly cite the
R package as 'Leo Lahti, Juuso Parkkinen ja Joona Lehtomäki (C)
2014. eurostat R package. URL: http://ropengov.github.io/eurostat'.


### Session info

This tutorial was created with

```{r sessioninfo, message=FALSE, warning=FALSE}
sessionInfo()
```
