<!--
%\VignetteEngine{knitr}
%\VignetteIndexEntry{An R Markdown Vignette made with knitr}
-->

Eurostat R tools
===========

This R package provides tools to access open data from [Eurostat](http://epp.eurostat.ec.europa.eu/portal/page/portal/statistics/themes). 

This R package is part of the [rOpenGov](http://ropengov.github.io)
project.


## Installation

Release version for general use:

```{r install, eval=FALSE}
install.packages("eurostat")
library(eurostat)
```

Development version (potentially unstable):

```{r install2, eval=FALSE}
install.packages("devtools")
library(devtools)
install_github("eurostat", "ropengov")
library(eurostat)
```


## Accessing Eurostat data from Eurostat database

```{r eurostat, warning=FALSE, message=FALSE}
library(eurostat)

# info about passagers
grepEurostatTOC("split of passenger transport")

## get table
tmp <- getEurostatRCV("tsdtr210")
summary(tmp)
```




## Accessing Eurostat data from Statistics Finland 

Eurostat data is also available through the [Statistics
Finland](http://www.stat.fi/tup/tilastotietokannat/index_fi.html) data
portal, where the listings of
[Eurostat](http://www.stat.fi/org/lainsaadanto/avoin_data.html) data
sets are available for browsing in
[PCAxis](http://pxweb2.stat.fi/Database/Eurostat/databasetree_fi.asp)
and [CSV](http://pxweb2.stat.fi/database/Eurostatn/Eurostatn_rap.csv)
formats. You can retrieve Eurostat data from Statfi by defining the
URL of the selected data set:

```{r tilastokeskus-2, warning=FALSE, message=FALSE}
# Define URL (browse the statfi site for URL listing; see above)
url <- "http://pxweb2.stat.fi/Database/Eurostat/ymp/t2020_30.px"

# Download the data
df <- get_eurostat(url)

# Inspect the first entries..
df[1:3,]
```


## Related tools

This R package is based on two earlier CRAN packages: statfi and smarterpoland. The [datamart](http://cran.r-project.org/web/packages/datamart/index.html) package contains additional related tools for Eurostat data but at the time of writing this tutorial this package seems to be in an experimental stage.


## Licensing and Citations

### Citing the Data

Regarding the data, kindly cite [Eurostat](http://epp.eurostat.ec.europa.eu/portal/page/portal/statistics/search_database). 


### Citing the R tools

This work can be freely used, modified and distributed under the
[GPL-3 license](https://www.gnu.org/copyleft/gpl.html). Kindly cite
the R package as 'Leo Lahti and Przemyslaw Biecek (C) 2014. eurostat R
package. URL: http://ropengov.github.io/eurostat'.


### Session info

This tutorial was created with

```{r sessioninfo, message=FALSE, warning=FALSE}
sessionInfo()
```
