<!--
%\VignetteEngine{knitr}
%\VignetteIndexEntry{An R Markdown Vignette made with knitr}
-->

Eurostat R tools
===========

This R package provides tools to access open data from [Eurostat](http://epp.eurostat.ec.europa.eu/portal/page/portal/statistics/themes). 

The present implementation retrieves the Eurostat data through the [Statistics
Finland](http://www.stat.fi/tup/tilastotietokannat/index_fi.html) data portal. Pull requests for extensions are welcome.

This R package is part of the [rOpenGov](http://ropengov.github.io)
project.

The listings of [Eurostat](http://www.stat.fi/org/lainsaadanto/avoin_data.html) data sets are available for browsing in [PCAxis](http://pxweb2.stat.fi/Database/Eurostat/databasetree_fi.asp) and [CSV](http://pxweb2.stat.fi/database/Eurostatn/Eurostatn_rap.csv) formats.


## Installation

Release version for general use:

```{r install, eval=FALSE}
install.packages("eurostat")
library(eurostat)
```

Development version (potentially unstable):

```{r install2, eval=FALSE}
install.packages("devtools")
library(devtools)
install_github("eurostat", "ropengov")
library(eurostat)
```


### Browsing the data

List of the data sets is available in R:

```{r tilastokeskus-1, warning=FALSE, message=FALSE}
# Load the library
library(eurostat)

# Eurostat open data listing
datasets.eurostat <- list_eurostat_files()

# Descriptions of the first entries
head(datasets.eurostat$DESCRIPTION)

# Investigate the first entry 
print(datasets.eurostat[1,])
```

## Retrieving the data

Retrieve data from Eurostat by defining URL of the data set. For the
listing of available data sets and their corresponding URLs, see
above.

```{r tilastokeskus-2, warning=FALSE, message=FALSE}
library(eurostat)

# Define URL (see list_eurostat_files())
url <- "http://pxweb2.stat.fi/Database/Eurostat/ymp/t2020_30.px"

# Download the data
df <- get_eurostat(url)

# Inspect the first entries..
df[1:3,]
```


## Related tools

http://www.inside-r.org/packages/cran/datamart/docs/eurostat

http://stackoverflow.com/questions/12762431/is-this-the-solution-to-get-data-from-eurostat-into-r

http://r.789695.n4.nabble.com/Download-CSV-Files-from-EUROSTAT-Website-td4679464.html



## Licensing and Citations

### Citing the Data

Regarding the data, kindly cite [Eurostat](http://epp.eurostat.ec.europa.eu/portal/page/portal/statistics/search_database). 


### Citing the R tools

This work can be freely used, modified and distributed under the
[Two-clause FreeBSD
license](http://en.wikipedia.org/wiki/BSD\_licenses). Kindly cite the
R package as 'Leo Lahti, Juuso Parkkinen ja Joona Lehtomäki (C)
2014. eurostat R package. URL: http://ropengov.github.io/eurostat'.


### Session info

This tutorial was created with

```{r sessioninfo, message=FALSE, warning=FALSE}
sessionInfo()
```
