<!--
%\VignetteIndexEntry{eurostat tutorial}
%\VignetteEngine{knitr::knitr}
%\usepackage[utf8]{inputenc}
-->    



Eurostat R tools
===========

This R package provides tools to access 
[Eurostat database](http://ec.europa.eu/eurostat/) 
as part of the [rOpenGov](http://ropengov.github.io) project.

For contact information and source code, see the [github page](https://github.com/rOpenGov/eurostat)

## Available tools

 * [Installation](#installation)
 * [Finding data](#search)
 * [Downloading data](#download)  
 * [Replacing codes with labels](#labeling)  
 * [Selecting and modifying data](#select)  
 * [Visualization](#visualization)  
   * [Triangle plot](#triangle)  
 * [Maps](#maps)  
 * [Citing the package](#citing)  
 * [Acknowledgements](#acknowledgements)  
 * [Session info](#session)  


```{r setup, include=FALSE}
# Global options
library(knitr)
opts_chunk$set(fig.path="fig/")
```

## <a name="installation"></a>Installation

Release version:

```{r install, eval=FALSE}
install.packages("eurostat")
```

Development version:

```{r install2, eval=FALSE}
library(devtools)
install_github("ropengov/eurostat")
```

Overall, the eurostat package includes the following functions:

```{r, echo=TRUE}
library(eurostat)
eurostat.functions <- sort(ls("package:eurostat"))
kable(as.data.frame(eurostat.functions))
```


## <a name="search"></a>Finding data

Function `get_eurostat_toc()` downloads a table of contents of eurostat datasets. The values in column 'code' should be used to download a selected dataset.

```{r get_eurostat_toc, warning=FALSE, message=FALSE}
# Load the package
library(eurostat)
library(rvest)

# Get Eurostat data listing
toc <- get_eurostat_toc()

# Check the first items
library(knitr)
kable(head(toc))
```

With `search_eurostat()` you can search the table of contents for particular patterns, e.g. all datasets related to *passenger transport*. The kable function to produces nice markdown output. Note that with the `type` argument of this function you could restrict the search to for instance datasets or tables.

```{r search_eurostat, warning=FALSE, message=FALSE}
# info about passengers
kable(head(search_eurostat("passenger transport")))
```

Codes for the dataset can be searched also from the [Eurostat
database](http://ec.europa.eu/eurostat/data/database). The Eurostat
database gives codes in the Data Navigation Tree after every dataset
in parenthesis.

## <a name="download"></a>Downloading data 

The package supports two of the Eurostats download methods: the bulk download 
facility and the Web Services' JSON API. The bulk download facility is the 
fastest method to download whole datasets. It is also often the only way as 
the JSON API has limitation of maximum 50 sub-indicators at a time and 
whole datasets usually exceeds that. To download only a small section of the 
dataset the JSON API is faster, as it allows to make a data selection 
before downloading.

A user does not usually have to bother with methods, as both are used via main
function `get_eurostat()`. If only the table id is given, the whole table is 
downloaded from the bulk download facility. If also filters are defined 
the JSON API is used.

Here an example of indicator [Modal split of passenger transport](http://ec.europa.eu/eurostat/tgm/table.do?tab=table&init=1&plugin=1&language=en&pcode=tsdtr210). This is the percentage share of each mode of transport in total inland transport, expressed in passenger-kilometres (pkm) based on transport by passenger cars, buses and coaches, and trains. All data should be based on movements on national territory, regardless of the nationality of the vehicle. However, the data collection is not harmonized at the EU level. 

Pick and print the id of the data set to download: 
```{r get_id, warning=FALSE, message=FALSE, results='asis'}
id <- search_eurostat("Modal split of passenger transport", 
        	             type = "table")$code[1]
print(id)
```

Get the whole corresponding table. As the table is annual data, it is more
convient to use a numeric time variable than use the default date format:

```{r get_eurostat, warning=FALSE, message=FALSE, results='asis'}
dat <- get_eurostat(id, time_format = "num")
```

Investigate the structure of the downloaded data set:
```{r str_dat, warning=FALSE, message=FALSE}
str(dat)
```

```{r head_dat, warning=FALSE, message=FALSE, results='asis'}
kable(head(dat))
```

Or you can get only a part of the dataset by defining `filters` argument. It
should be named list, where names corresponds to variable names (lower case) and
values are vectors of codes corresponding desidered series (upper case). For
time variable, in addition to a `time`, also a `sinceTimePeriod` 
and a `lastTimePeriod` can be used.

```{r get_eurostat_json, warning=FALSE, message=FALSE, results='asis'}

dat2 <- get_eurostat(id, filters = list(geo = c("EU28", "FI"), lastTimePeriod=1), time_format = "num")
kable(dat2)
```

### <a name="labeling"></a>Replacing codes with labels

By default variables are returned as Eurostat codes, but to get human-readable 
labels instead, use a `type = "label"` argument.  

```{r json_labels, warning=FALSE, message=FALSE, results='asis'}
datl2 <- get_eurostat(id, filters = list(geo = c("EU28", "FI"), 
                                         lastTimePeriod = 1), 
                      type = "label", time_format = "num")
kable(head(datl2))
```

Eurostat codes can be replaced also after downloadind with human-readable labels
using a function `label_eurostat()`. It replaces the eurostat codes based on
definitions from Eurostat dictionaries. 

```{r labels, warning=FALSE, message=FALSE, results='asis'}
datl <- label_eurostat(dat)
kable(head(datl))
```

The `label_eurostat()` allows also 
conversion of individual variable vectors or variable names. 

```{r name_labels}

label_eurostat_vars(names(datl))
```


Vehicle information has 3 levels. They are:
```{r vehicle_levels}
levels(datl$vehicle)
```



## <a name="select"></a>Selecting and modifying data

### EFTA, Eurozone, EU and EU candidate countries

To facilititate fast plotting of standard European geographic areas, the package provides ready-made lists of the country codes used in the eurostat database for EFTA (efta\_countries), Euro area (ea\_countries), EU (eu\_countries) and EU candidate countries (candidate\_countries). This helps to select specific groups of countries for closer investigation. For conversions with other standard country coding systems, see the [countrycode](...) R package. To retrieve the country code list for EFTA, for instance, use:

```{r, echo=TRUE, eval=TRUE}
data(efta_countries)
kable(efta_countries)
```


### EU data from 2012 in all vehicles:

```{r eu_12}
dat_eu12 <- subset(datl, geo == "European Union (28 countries)" & time == 2012)
kable(dat_eu12, row.names = FALSE)
```

### EU data from 2000 - 2012 with vehicle types as variables:

Reshaping the data is best done with `spread()` in `tidyr`.
```{r eu_vehicles_table}
library("tidyr")
dat_eu_0012 <- subset(dat, geo == "EU28" & time %in% 2000:2012)
dat_eu_0012_wide <- spread(dat_eu_0012, vehicle, values)
kable(subset(dat_eu_0012_wide, select = -geo), row.names = FALSE)
```

### Train passengers for selected EU countries in 2000 - 2012

```{r trains_table}
dat_trains <- subset(datl, geo %in% c("Austria", "Belgium", "Finland", "Sweden")
                     & time %in% 2000:2012 
                     & vehicle == "Trains")

dat_trains_wide <- spread(dat_trains, geo, values) 
kable(subset(dat_trains_wide, select = -vehicle), row.names = FALSE)
```



## <a name="visualization"></a>Visualization

Visualizing train passenger data with `ggplot2`:

```{r trains_plot, fig.width=8, fig.height=3}
library(ggplot2)
p <- ggplot(dat_trains, aes(x = time, y = values, colour = geo)) 
p <- p + geom_line()
print(p)
```

<a name="triangle"></a>**Triangle plot**

Triangle plot on passenger transport distributions with 2012 data for all countries with data.

```{r plotGallery, warning=FALSE, message=FALSE}
library(tidyr)

transports <- spread(subset(dat, time == 2012, select = c(geo, vehicle, values)), vehicle, values)

transports <- na.omit(transports)

# triangle plot
library(plotrix)
triax.plot(transports[, -1], show.grid = TRUE, 
           label.points = TRUE, point.labels = transports$geo, 
           pch = 19)
```

## <a name="maps"></a>Maps 

**Quick demo1 - whole Europe at country level reso 1:60mln**

```{r maps7, eval=FALSE}
# Data from Eurostat
dat <- eurostat::get_eurostat(id = "tsdtr420", time_format = "num")
# subset to have only a single row per geo
dat <- dat %>% dplyr::filter(sex == "T", time %in% 2014)

# merge with geodata
ff <- merge_with_geodata(data=dat,geocol="geo",res = "60")

# plot map
ggplot(data=ff, 
       aes(x=long,y=lat,group=group)) +
  geom_polygon(aes(fill=values),color="white")
```

**Quick demo2 - Poland at NUTS3 level reso 1:1mln**

```{r maps8, eval=FALSE}
# Data from Eurosrat
dat <- eurostat::get_eurostat("tgs00026", time_format = "raw")
# subset to have only a single row per geo
dat <- dat %>% dplyr::filter(time == 2010, grepl("PL",geo), nchar(as.character(geo)) == 4)

# merge with geodata
ff <- merge_with_geodata(data=dat,geocol="geo",res = "01")

# plot map
ggplot(data=ff, 
       aes(x=long,y=lat,group=group)) +
  geom_polygon(aes(fill=values),color="white")
```

**Quick demo3 - whole Europe at country level reso 1:60mln**

```{r maps1, eval=FALSE}

# Data from Eurostat
dat <- eurostat::get_eurostat(id = "tsdtr420", time_format = "num")
# subset to have only a single row per geo
dat <- dat %>% dplyr::filter(sex == "T", time == 2014)

# categorise
dat$cat <- cut_to_classes(dat$values)

# merge with geodata
ff <- merge_with_geodata(data=dat,geocol="geo",res = "60")

# plot map
ggplot(data=ff, 
       aes(x=long,y=lat,group=group)) +
  geom_polygon(aes(fill=cat),color="white") +
  scale_fill_brewer(palette ="Oranges")
```

**Quick demo4 - Poland at NUTS3 level reso 1:1mln**

```{r maps2, eval=FALSE}
dat <- eurostat::get_eurostat("tgs00026", time_format = "raw")
# subset to have only a single row per geo
dat <- dat %>% dplyr::filter(time == 2010, grepl("PL",geo), nchar(as.character(geo)) == 4)

# categorise
dat$cat <- cut_to_classes(dat$values)

# merge with geodata
ff <- merge_with_geodata(data=dat,geocol="geo",res = "01")

# plot map
ggplot(data=ff, 
       aes(x=long,y=lat,group=group)) +
  geom_polygon(aes(fill=cat),color="white") +
  scale_fill_brewer(palette ="Oranges")
```

**Reproducing the map in the paper**

```{r maps4}
library(eurostat)
# Downloading and manipulating the tabular data
df <- get_eurostat("tgs00026", time_format = "raw")

# subsetting to year 2005 and NUTS-3 level
df <- df %>% dplyr::filter(time == 2005, nchar(as.character(geo)) == 4)

# categorise
df$cat <- cut_to_classes(df$values)

# merge with geodata
plot_map <- merge_with_geodata(data=df,geocol="geo",res = "60")

# plot map
p <- ggplot(data=plot_map, aes(long,lat,group=group))
p <- p + geom_polygon(aes(fill = cat),colour="white",size=.8)
p <- p + scale_fill_manual(values=c("dim grey",RColorBrewer::brewer.pal(n = 5, name = "Oranges"))) 
p <- p + coord_map(project="orthographic", xlim=c(-22,34), ylim=c(35,70))
p <- p + labs(title = paste0("Disposable household incomes in 2005"))
p <- p +  theme(legend.position = c(0.03,0.40), 
                legend.justification=c(0,0),
                legend.key.size=unit(6,'mm'),
                legend.direction = "vertical",
                legend.background=element_rect(colour=NA, fill=alpha("white", 2/3)),
                legend.text=element_text(size=12), 
                legend.title=element_text(size=12), 
                title=element_text(size=16), 
                panel.background = element_blank(), 
                plot.background = element_blank(),
                panel.grid.minor = element_line(colour = 'Grey80', size = .5, linetype = 'solid'),
                panel.grid.major = element_line(colour = 'Grey80', size = .5, linetype = 'solid'),
                axis.text = element_blank(), 
                axis.title = element_blank(), 
                axis.ticks = element_blank(), 
                plot.margin = unit(c(-3,-1.5, -3, -1.5), "cm"))
p <- p + guides(fill = guide_legend(title = "EUR per Year",
                                    title.position = "top", 
                                    title.hjust=0))
p
```


**Poland with labels**

```{r maps6}
library(eurostat)
# Downloading and manipulating the tabular data
df <- get_eurostat("tgs00026", time_format = "raw")

# subsetting to year 2005 and NUTS-3 level
df <- df %>% dplyr::filter(time == 2005, nchar(as.character(geo)) == 4, grepl("PL",geo))

df_lab <- label_eurostat(df)
names(df_lab) <- paste0("lab",names(df_lab))

df2 <- cbind(df,df_lab)

# categorise
df2$cat <- cut_to_classes(df2$values)

# merge with geodata
plot_map <- merge_with_geodata(data=df2,geocol="geo",res = "60")

cnames <- stats:::aggregate.formula(cbind(long, lat) ~ labgeo, data=plot_map, mean) # region names
cnames <- merge(cnames,df2,by="labgeo")

# plot map
p <- ggplot(data=plot_map, aes(long,lat,group=group))
p <- p + geom_polygon(aes(fill = cat),colour="white",size=.8)
p <- p + scale_fill_manual(values=RColorBrewer::brewer.pal(n = 5, name = "Oranges"))
p <- p + geom_label(data=cnames, aes(long, lat, label = paste(labgeo,"\n",values,"€"), group=labgeo,fill=cat), 
                    size=3.5, color="white", fontface="bold", lineheight=.8, show.legend=FALSE)
p <- p + coord_map(project="orthographic")
p <- p + labs(title = paste0("Disposable household incomes in 2005"))
p <- p +  theme(legend.position = c(0.1,0.03), 
                legend.justification=c(0,0),
                legend.key.size=unit(6,'mm'),
                legend.direction = "vertical",
                legend.background=element_rect(colour=NA, fill=alpha("white", 2/3)),
                legend.text=element_text(size=12), 
                legend.title=element_text(size=12), 
                title=element_text(size=16), 
                panel.background = element_blank(), 
                plot.background = element_blank(),
                panel.grid.minor = element_line(colour = 'Grey80', size = .5, linetype = 'solid'),
                panel.grid.major = element_line(colour = 'Grey80', size = .5, linetype = 'solid'),
                axis.text = element_blank(), 
                axis.title = element_blank(), 
                axis.ticks = element_blank(), 
                plot.margin = unit(c(-3,-1.5, -3, -1.5), "cm"))
p <- p + guides(fill = guide_legend(title = "EUR per Year",
                                    title.position = "top", 
                                    title.hjust=0))
p
```






For further examples, see also the [blog post on the eurostat R package](http://ropengov.github.io/r/2015/05/01/eurostat-package-examples/).


## <a name="citing"></a>Citing the package

**Citing the Data** Kindly cite [Eurostat](http://ec.europa.eu/eurostat/). 

**Citing the R tools** This work can be freely used, modified and
distributed under the BSD-2-clause (modified FreeBSD) license:

```{r citation, message=FALSE, eval=TRUE}
citation("eurostat")
```

## <a name="acknowledgements"></a>Acknowledgements


We are grateful to all [contributors](https://github.com/rOpenGov/eurostat/graphs/contributors) and [Eurostat](http://ec.europa.eu/eurostat/) open data portal! This [rOpenGov](http://ropengov.github.io) R package is based on earlier CRAN packages [statfi](https://cran.r-project.org/package=statfi) and [smarterpoland](https://cran.r-project.org/package=SmarterPoland). The [datamart](https://cran.r-project.org/package=datamart) and [reurostat](https://github.com/Tungurahua/reurostat) packages seem to develop related Eurostat tools but at the time of writing this tutorial this package seems to be in an experimental stage. The [quandl](https://cran.r-project.org/package=quandl) package may also provides access to some versions of eurostat data sets.


## <a name="session"></a>Session info

This tutorial was created with

```{r sessioninfo, message=FALSE, warning=FALSE}
sessionInfo()
```
