<!--
%\VignetteIndexEntry{eurostat tutorial}
%\VignetteEngine{knitr::knitr}
%\usepackage[utf8]{inputenc}
-->    



Eurostat R tools
===========

This R package provides tools to access 
[Eurostat database](http://ec.europa.eu/eurostat/) 
as part of the [rOpenGov](http://ropengov.github.io) project.

For contact information and source code, see the [github page](https://github.com/rOpenGov/eurostat)

## Available tools

 * [Installation](#installation)
 * [Finding data](#search)
 * [Downloading data](#download)  
 * [Replacing codes with labels](#labeling)  
 * [Selecting and modifying data](#select)  
 * [Visualization](#visualization)  
   * [Triangle plot](#triangle)  
 * [Citing the package](#citing)  
 * [Acknowledgements](#acknowledgements)  
 * [Session info](#session)  


```{r setup, include=FALSE}
# Global options
library(knitr)
opts_chunk$set(fig.path="fig/")
```

## <a name="installation"></a>Installation

Release version:

```{r install, eval=FALSE}
install.packages("eurostat")
```

Development version:

```{r install2, eval=FALSE}
library(devtools)
install_github("ropengov/eurostat")
```

Overall, the eurostat package includes the following functions:

```{r, echo=TRUE}
library(eurostat)
kable(as.data.frame(ls("package:eurostat")))
```


## <a name="search"></a>Finding data

Function `get_eurostat_toc()` downloads a table of contents of eurostat datasets. The values in column 'code' should be used to download a selected dataset.

```{r get_eurostat_toc, warning=FALSE, message=FALSE}
# Load the package
library(eurostat)
library(rvest)

# Get Eurostat data listing
toc <- get_eurostat_toc()

# Check the first items
library(knitr)
kable(head(toc))
```

With `search_eurostat()` you can search the table of contents for particular patterns, e.g. all datasets related to *passenger transport*. The kable function to produces nice markdown output. Note that with the `type` argument of this function you could restrict the search to for instance datasets or tables.

```{r search_eurostat, warning=FALSE, message=FALSE}
# info about passengers
kable(head(search_eurostat("passenger transport")))
```

Codes for the dataset can be searched also from the [Eurostat
database](http://ec.europa.eu/eurostat/data/database). The Eurostat
database gives codes in the Data Navigation Tree after every dataset
in parenthesis.

## <a name="download"></a>Downloading data 

Here an example of indicator [Modal split of passenger transport](http://ec.europa.eu/eurostat/tgm/table.do?tab=table&init=1&plugin=1&language=en&pcode=tsdtr210). This is the percentage share of each mode of transport in total inland transport, expressed in passenger-kilometres (pkm) based on transport by passenger cars, buses and coaches, and trains. All data should be based on movements on national territory, regardless of the nationality of the vehicle. However, the data collection is not harmonized at the EU level. 

Pick and print the id of the data set to download: 
```{r get_id, warning=FALSE, message=FALSE, results='asis'}
id <- search_eurostat("Modal split of passenger transport", 
        	             type = "table")$code[1]
print(id)
```

Get the corersponding table. As the table is annual data, it is more
convient to use a numeric time variable than use the default date format:

```{r get_eurostat, warning=FALSE, message=FALSE, results='asis'}
dat <- get_eurostat(id, time_format = "num")
```

Investigate the structure of the downloaded data set:
```{r str_dat, warning=FALSE, message=FALSE, results='asis'}
str(dat)
kable(head(dat))
```

### <a name="labeling"></a>Replacing codes with labels

Eurostat variable IDs can be replaced with human-readable labels.
Function `label_eurostat()` replaces the eurostat IDs based on
definitions from Eurostat dictionaries.

```{r labels, warning=FALSE, message=FALSE, results='asis'}
datl <- label_eurostat(dat)
kable(head(datl))
```

Vehicle information has 3 levels. They are:
```{r vehicle_levels}
levels(datl$vehicle)
```

You can get also labels for variable names
```{r name_labels}

label_eurostat_vars(names(datl))
```


## <a name="select"></a>Selecting and modifying data

### EFTA, Eurozone, EU and EU candidate countries

To facilititate fast plotting of standard European geographic areas, the package provides ready-made lists of the country codes used in the eurostat database for EFTA (efta\_countries), Euro area (ea\_countries), EU (eu\_countries) and EU candidate countries (candidate\_countries). This helps to select specific groups of countries for closer investigation. For conversions with other standard country coding systems, see the [countrycode](...) R package. To retrieve the country code list for EFTA, for instance, use:

```{r, echo=TRUE, eval=TRUE}
data(efta_countries)
print(efta_countries)
```


### EU data from 2012 in all vehicles:

```{r eu_12}
dat_eu12 <- subset(datl, geo == "European Union (28 countries)" & time == 2012)
kable(dat_eu12, row.names = FALSE)
```

### EU data from 2000 - 2012 with vehicle types as variables:

Reshaping the data is best done with `spread()` in `tidyr`.
```{r eu_vehicles_table}
library("tidyr")
dat_eu_0012 <- subset(dat, geo == "EU28" & time %in% 2000:2012)
dat_eu_0012_wide <- spread(dat_eu_0012, vehicle, values)
kable(subset(dat_eu_0012_wide, select = -geo), row.names = FALSE)
```

### Train passengers for selected EU countries in 2000 - 2012

```{r trains_table}
dat_trains <- subset(datl, geo %in% c("Austria", "Belgium", "Finland", "Sweden")
                     & time %in% 2000:2012 
                     & vehicle == "Trains")

dat_trains_wide <- spread(dat_trains, geo, values) 
kable(subset(dat_trains_wide, select = -vehicle), row.names = FALSE)
```



## <a name="visualization"></a>Visualization

Visualizing train passenger data with `ggplot2`:

```{r trains_plot, fig.width=10, fig.height=4}
library(ggplot2)
p <- ggplot(dat_trains, aes(x = time, y = values, colour = geo)) 
p <- p + geom_line()
print(p)
```

### <a name="triangle"></a>Triangle plot

Triangle plot on passenger transport distributions with 2012 data for all countries with data.

```{r plotGallery, warning=FALSE, message=FALSE}
library(tidyr)

transports <- spread(subset(dat, time == 2012, select = c(geo, vehicle, values)), vehicle, values)

transports <- na.omit(transports)

# triangle plot
library(plotrix)
triax.plot(transports[, -1], show.grid = TRUE, 
           label.points = TRUE, point.labels = transports$geo, 
           pch = 19)
```

For further examples, see also the [blog post on the eurostat R
package](...).


## <a name="citing"></a>Citing the package

**Citing the Data** Kindly cite [Eurostat](http://ec.europa.eu/eurostat/). 

**Citing the R tools** This work can be freely used, modified and
distributed under the BSD-2-clause (modified FreeBSD) license:

```{r citation, message=FALSE, eval=TRUE}
citation("eurostat")
```

## <a name="acknowledgements"></a>Acknowledgements


We are grateful to all [contributors](https://github.com/rOpenGov/eurostat/graphs/contributors) and [Eurostat](http://ec.europa.eu/eurostat/) open data portal! This [rOpenGov](http://ropengov.github.io) R package is based on earlier CRAN packages [statfi](https://cran.r-project.org/package=statfi) and [smarterpoland](https://cran.r-project.org/package=SmarterPoland). The [datamart](https://cran.r-project.org/package=datamart) and [reurostat](https://github.com/Tungurahua/reurostat) packages seem to develop related Eurostat tools but at the time of writing this tutorial this package seems to be in an experimental stage. The [quandl](https://cran.r-project.org/package=quandl) package may also provides access to some versions of eurostat data sets.


## <a name="session"></a>Session info

This tutorial was created with

```{r sessioninfo, message=FALSE, warning=FALSE}
sessionInfo()
```
