---
title: "eurostat R package"
author: Przemyslaw Biecek, Janne Huovari, Markus Kainu, Leo Lahti
date: "`r Sys.Date()`"
bibliography: 
- bibliography.bib
- references.bib
output: 
  md_document:
    variant: markdown_github
---
<!--
%\VignetteEngine{knitr::rmarkdown}
%\VignetteIndexEntry{eurostat Markdown Vignette}
%\usepackage[utf8]{inputenc}
-->


```{r 2015-manu-init0, echo=FALSE, message=FALSE}
# Handle citations
require(knitcitations)
cleanbib()
#options("citation_format" = "pandoc")
bib <- read.bibtex("bibliography.bib")
#bib2 <- read.bibtex("references.bib")
#opts_chunk$set(fig.width=4, fig.height=3, par=TRUE, out.width='2in', fig.pos='H')

# figure folder
knitr::opts_chunk$set(fig.path = "2015-manuscript-figures/")

# ggplot theme
theme_set(theme_bw(20))
```

```{r 2015-manu-opts, echo=FALSE}
library(eurostat)
library(knitr)
opts_chunk$set(list(echo=TRUE,eval=TRUE,cache=FALSE,warning=FALSE,message=FALSE,fig.height=10,fig.width=10))
```

```{r 2015-manu-functions, include=FALSE}
# One way for automated figure numbering and citing, borrowed from
# http://stackoverflow.com/questions/13848137/figure-captions-references-using-knitr-and-markdown-to-html

# A function for captioning and referencing images
fig <- local({
    i <- 0
    ref <- list()
    list(
        cap=function(refName, text) {
            i <<- i + 1
            ref[[refName]] <<- i
            paste("Figure ", i, ": ", text, sep="")
        },
        ref=function(refName) {
            ref[[refName]]
        })
})
```

To be submitted to [R Journal](https://journal.r-project.org/) ?

## Introduction

Governmental institutions have started to release increasing amount of
their data resources for the public as open data in the recent
years. This is opening novel opportunities for research and citizen
science. Eurostat, for instance, provides a rich collection of
demographic and economic data through its open data portal. The portal
currently includes ... data sets on ... between years ...

Efficient tools to access and analyse such data collections can
greatly benefit reproducible research. When the data is available, the
analytical methodology spanning from raw data to the final publication
can also be made available following reproducible research principles
`r citep(bib[["Gandrud13"]])`. Standardization and automatization of
common data analysis tasks via dedicated software packages can greatly
facilitate reproducibility and code sharing, making the research more
transparent and efficient.

Here, we introduce the eurostat R package that implements automated
tools for R to access [open data from
Eurostat](http://ec.europa.eu/eurostat/) in [R statistical programming
language](http://www.r-project.org) `r citep(bib[["rcore13"]])`. The
package has been actively developed by several independent
contributors and based on the community feedback in Github, and its
first CRAN release was on.. We are now reporting the first mature
version of the package that has been improved and tested by multiple
users, and includes features like cache, handling dates, and using the
tidy data principles `r citep(bib[["tidydata"]])`, with the help of
the tidyr R package `r citep(citation("tidyr"))`.


The work is based on our earlier CRAN packages
[statfi](http://cran.r-project.org/web/packages/statfi/index.html) and
[smarterpoland](http://cran.r-project.org/web/packages/SmarterPoland/index.html). Compared to this earlier work, we have now implemented an expanded
set of tools specifically focusing on the eurostat data collection. The
[datamart](http://cran.r-project.org/web/packages/datamart/index.html)
and the
[quandl](http://cran.r-project.org/web/packages/Quandl/index.html) R
packages [ADD REFS!!!] provide also access to certain versions of
eurostat data. In contrast to these generic database packages, our
work is fully focused on the Eurostat open data portal and provides
specific functionality suited for this data collection. There is a
development version for R package
[reurostat](https://github.com/Tungurahua/reurostat) but it does not
seem to be actively maintained at the moment (last commits at
...). The eurostat R package takes advantage of the
following external R packages:
devtools `r citep(citation("devtools"))`,
dplyr `r citep(citation("dplyr"))`,
knitr `r citep(citation("knitr"))`,
ggplot2 `r citep(citation("ggplot2"))`,
mapproj `r citep(citation("mapproj"))`,
plotrix `r citep(citation("plotrix"))`,
reshape2 `r citep(citation("reshape2"))`,
rmarkdown `r citep(citation("rmarkdown"))`,
stringi `r citep(citation("stringi"))`,
testthat `r citep(citation("testthat"))`,
tidyr `r citep(citation("tidyr"))`.
The eurostat R package is part of the
[rOpenGov](http://ropengov.github.io) collection
`r citep(bib[["Lahti13icml"]])`, which provides reproducible research
tools for computational social science and digital humanities.





## Overview of the functionality

The package includes tools to search and retrieve specific data sets
from the Eurostat open data portal, converting identifiers in
human-readable formats, selecting, modifying and visualizing the
data. Detailed installation instructions and usage examples are
provided in the [package
vignette](https://github.com/rOpenGov/eurostat/vignette/eurostat_tutorial.Rmd), and the [introductory blog post](http://ropengov.github.io/r/2015/05/01/eurostat-package-examples).

The unified interface to data sets can make data analysis more
straightforward and transparent by providing a standardized and
automated way to access the data sets. Here we describe the
functionality of the current CRAN release version (1.2.1). To install
this, simply use

```{r 2015-manu-install, eval=FALSE}
install.packages("eurostat")
```

You can download the complete table of contents of the database with the function `get_eurostat_toc()`, or use `search_eurostat()` to make a more focused search over the table of contents. To retrieve data sets for 'disposable income', for instance, use:


```{r 2015-manu-init, eval=TRUE}
library(eurostat)
results <- search_eurostat("disposable income", type = "dataset")
kable(head(results))
```

The data type to search for is also specified. The options include 'table', 'dataset' or 'folder', referring to different levels of hierarchy in data organization: a 'table' resides in 'dataset' that are stored in a 'folder'. You can focus the search on a selected type.

The first lines of the output are shown above. Use values from the `code` column to refer to a specific data set in the subsequent download commands. The dataset identifier codes can also be browsed at the Eurostat
database \url{http://ec.europa.eu/eurostat/data/database}, which gives
codes in the Data Navigation Tree after every dataset in parenthesis.

Download the selected data with `get_eurostat()`. To retrieve the data set with a particular identifier, use 

```{r 2015-manu-get_eurostat, warning=FALSE, message=FALSE, results='asis', eval=TRUE}
dat <- get_eurostat(id, time_format = "num")
```

The original data is annual in this example, hence we have here
selected a numeric time variable as it is more convient for annual
time series than the default date format. To improve the
interpretability of the output, the eurostat variable identifiers
could be further replaced with human-readable labels based on
definitions from Eurostat dictionaries with the `label_eurostat()`
function. The data is provided in the standard data.frame format, and
all standard tools for data subsetting and reshaping can be
conveniently applied.

The downloaded data sets are stored in cache by default. This can help
to avoid repeated downloading of identical data and helps to speed up
the analysis. Another advantage is that by storing an exact copy of
the data on the hard disk, it is possible to reproduce the analysis
results afterwards even if the source database has been updated.

Data sets containing geographic information can be visualized on a
map. Most indicators are of country-year -type, although some
indicators have data also at lower level of regional breakdown. The
disposable income of private households at [NUTS2](http://en.wikipedia.org/wiki/Nomenclature_of_Territorial_Units_for_Statistics) level can be
visualized, for instance, as in Figure XXX [PLACE THE FIGURE FROM MARKUS
BLOG POST WITH HIDDEN SOURCE CODE HERE]. For a more detailed treatment
of this example, see our related blog post [ADD LINK]. See more information on regional classification [here](http://ec.europa.eu/eurostat/ramon/nomenclatures/index.cfm?TargetUrl=DSP_GEN_DESC_VIEW_NOHDR&StrNom=NUTS_33&StrLanguageCode=EN).

Here, we have combined the data retrieved with the eurostat package with additional map visualization tools and utilities including
grid `r citep(citation("grid"))`,
maptools `r citep(citation("maptools"))`,
rgdal `r citep(citation("rgdal"))`,
rgeos `r citep(citation("rgeos"))`, 
scales `r citep(citation("scales"))`, and
stringr `r citep(citation("stringr"))`.

Example on spatio-temporal data visualization: let us look at the indicator [`tgs00026`](http://ec.europa.eu/eurostat/en/web/products-datasets/-/TGS00026), (*Disposable income of private households by NUTS 2 regions*) from Eurostat. We are looking at the disposable household income. In addition to downloading and manipulating data from EUROSTAT, we demonstrate how to access and use shapefiles of Europe published by EUROSTAT at [Administrative units / Statistical units](http://ec.europa.eu/eurostat/web/gisco/geodata/reference-data/administrative-units-statistical-units).


For the map example, look at Figure `r fig$ref("2015-manu-RJournal_map_example")`.

```{r 2015-manu-RJournal_map_example, warning=FALSE, message=FALSE, echo=FALSE, eval=TRUE, fig.cap=fig$cap("2015-manu-RJournal_map_example", "Eurostat spatial visualization example.")}
# Clean the cache
clean_eurostat_cache()

# Load dependencies
PACKAGES <- c("rgdal","maptools","rgeos","stringr","scales","grid")
lapply(PACKAGES, require, character.only=T)

# Downloading and manipulating the tabular data
df <- get_eurostat("tgs00026", time_format = "raw")

# convert time column from Date to numeric
df$time <- eurotime2num(df$time)

# subset time to have data for 
df <- df[df$time %in% c(2005,2010,2011),]

# spread the data into wide format
library(tidyr)
dw <- spread(df, time, values)

# Downloading and manipulating the spatial data
# We download the zipped shapefile in 1:60 million scale from year
# 2010 and subset it at the level of NUTS2.
# Load the GISCO shapefile
download.file("http://ec.europa.eu/eurostat/cache/GISCO/geodatafiles/NUTS_2010_60M_SH.zip", destfile="NUTS_2010_60M_SH.zip")
# unzip
unzip("NUTS_2010_60M_SH.zip")

# read into SpatialPolygonsDataFrame
library(rgdal)
map <- readOGR(dsn = "./NUTS_2010_60M_SH/Data", layer = "NUTS_RG_60M_2010")

# subset the spatialpolygondataframe at NUTS2-level
map_nuts2 <- subset(map, STAT_LEVL_ == 2)

# Joining tabular data with spatial data
# Third, we will make the both datas of same length,
# give then identical rownames and then merge the tabular data with the
# spatial data.

# Spatial dataframe has 467 rows and attribute data 275.
# We need to make attribute data to have similar number of rows
NUTS_ID <- as.character(map_nuts2$NUTS_ID)
VarX <- rep(NA, 316)
dat <- data.frame(NUTS_ID,VarX)

# then we shall merge this with Eurostat data.frame
dat2 <- merge(dat, dw, by.x="NUTS_ID", by.y="geo", all.x=TRUE)

# merge this manipulated attribute data with the spatialpolygondataframe
# rownames
row.names(dat2) <- dat2$NUTS_ID
row.names(map_nuts2) <- as.character(map_nuts2$NUTS_ID)

# order data
dat2 <- dat2[order(row.names(dat2)), ]
map_nuts2 <- map_nuts2[order(row.names(map_nuts2)), ]

# join
library(maptools)
dat2$NUTS_ID <- NULL
shape <- spCbind(map_nuts2, dat2)

# Preparing the data for ggplot2 visualization
## fortify spatialpolygondataframe into data.frame
library(ggplot2)
library(rgeos)
shape$id <- rownames(shape@data)
map.points <- fortify(shape, region = "id")
map.df <- merge(map.points, shape, by = "id")

# As we want to plot map faceted by years from 2003 to 2011
# we have to melt it into long format
# (variable with numerical names got X-prefix during the spCbind-merge,
# therefore the X-prefix in variable names)
library(tidyr)

# lets convert unit variable (that is a list) into character
map.df$unit <- as.character(map.df$unit)
map.df.l <- gather(map.df, "year", "value", 15:17)

# year variable (variable) is class string and type X20xx.
# Lets remove the X and convert it to numerical
library(stringr)
map.df.l$year <- str_replace_all(map.df.l$year, "X","")
map.df.l$year <- factor(map.df.l$year)
map.df.l$year <- as.numeric(levels(map.df.l$year))[map.df.l$year]

# Plotting the maps using ggplot2
library(ggplot2)
library(scales)
library(grid)

# Creating a custom function for creating the breaks and makeing them look neat
categories <- function(x, cat = 5) {
  
  library(stringr)
  levs <- as.data.frame(as.character(levels(cut_interval(x, cat))))
  names(levs) <- "orig"
  levs$mod <- str_replace_all(levs$orig, "\\[", "")
  levs$mod <- str_replace_all(levs$mod, "\\]", "")
  levs$mod <- str_replace_all(levs$mod, "\\(", "")
  levs$lower <- gsub(",.*$","", levs$mod)
  levs$upper <- gsub(".*,","", levs$mod)
  
  levs$lower <- factor(levs$lower)
  levs$lower <- round(as.numeric(levels(levs$lower))[levs$lower],0)
  
  levs$upper <- factor(levs$upper)
  levs$upper <- round(as.numeric(levels(levs$upper))[levs$upper],0)
  
  levs$labs <- paste(levs$lower,levs$upper, sep=" - ")
  
  labs <- as.character(c(levs$labs))
  y <- cut_interval(x, cat, right = FALSE, labels = labs)
  y <- as.character(y)
  y[is.na(y)] <- "No Data"
  y <- factor(y, levels=c("No Data",labs[1:cat]))
}

# years for for loop - now show just one year
years <- unique(map.df.l$year)[[1]]

# Loop over the three years
for (year in years) {
  
  # subset data
  plot_map <- map.df.l[map.df.l$year == year,]
  # set the breaks
  plot_map$value_cat <- categories(plot_map$value)
  
  p <- ggplot(data=plot_map, aes(long,lat,group=group))
  p <- p + geom_polygon(data = map.df.l, aes(long,lat),fill=NA,colour="white",size = 1)
  p <- p + geom_polygon(aes(fill = value_cat),colour="white",size=.2)
  p <- p + scale_fill_manual(values=c("Dim Grey","#d7191c","#fdae61","#ffffbf","#a6d96a","#1a9641")) 
  p <- p + coord_map(project="orthographic", xlim=c(-22,34), ylim=c(35,70))
  p <- p + labs(title = paste0("Disposable household incomes in  ",year))
  p <- p +  theme(legend.position = c(0.03,0.40), 
                          legend.justification=c(0,0),
                          legend.key.size=unit(6,'mm'),
                          legend.direction = "vertical",
                          legend.background=element_rect(colour=NA, fill=alpha("white", 2/3)),
                          legend.text=element_text(size=12), 
                          legend.title=element_text(size=12), 
                          title=element_text(size=16), 
                          panel.background = element_blank(), 
                          plot.background = element_blank(),
                          panel.grid.minor = element_line(colour = 'Grey80', size = .5, linetype = 'solid'),
                          panel.grid.major = element_line(colour = 'Grey80', size = .5, linetype = 'solid'),
                          axis.text = element_blank(), 
                          axis.title = element_blank(), 
                          axis.ticks = element_blank(), 
                          plot.margin = unit(c(-3,-1.5, -3, -1.5), "cm"))
  p <- p + guides(fill = guide_legend(title = "€ per Year",
                                     title.position = "top", 
                                     title.hjust=0))
  print(p)
}
```


To make a triangle map from the plotrix `r citep(citation("plotrix"))`
package provides an example on visualizing passenger transport data
distributions (Figure YYY):

```{r 2015-manu-searchdata, echo=FALSE, eval=TRUE, fig.width=8, fig.height=5, fig.cap=fig$cap("2015-manu-searchdata", "Triangle plot example.")}
id <- search_eurostat("Modal split of passenger transport", 
        	             type = "table")$code[1]

dat <- get_eurostat(id, time_format = "num")

# Get labels for variable names
datl <- label_eurostat(dat)
label_eurostat_vars(names(datl))
dat_trains <- subset(datl, geo %in% c("Austria", "Belgium", "Finland", "Sweden")
                     & time %in% 2000:2012 
                     & vehicle == "Trains")

# Visualizing train passenger data
library(ggplot2)
p <- ggplot(dat_trains, aes(x = time, y = values, colour = geo)) 
p <- p + geom_line()
print(p)

# Triangle plot on passenger transport distributions with 2012 data for
# all countries with data 
library(tidyr)
transports <- spread(subset(dat, time == 2012,
	   select = c(geo, vehicle, values)), vehicle, values)

# triangle plot
library(plotrix)
triax.plot(na.omit(transports)[, -1], show.grid = TRUE, 
           label.points = TRUE, point.labels = transports$geo, 
           pch = 19)
```

To facilitate fast plotting of standard European geographic areas, the package provides ready-made lists of the country codes used in the eurostat database for EFTA (efta\_countries), Euro area (ea\_countries), EU (eu\_countries) and EU candidate countries (candidate\_countries). This helps to select specific groups of countries for closer investigation. For conversions with other standard country coding systems, see the countrycode R package `r citep(citation("countrycode"))`. To retrieve the eurostat country code list for EFTA, for instance, use:

```{r 2015-manu-efta, echo=TRUE, eval=TRUE}
data(efta_countries)
kable(efta_countries)
```

## Applications

The package or its predecessors have already been applied in several case studies by us and independent developers. Financial Times, for instance, [have used R to access data from Eurostat](http://blog.revolutionanalytics.com/2015/04/financial-times-tracks-unemployment-with-r.html) using functions from SmarterPoland, the basis for our revised and expanded eurostat package.

The [archivist](http://pbiecek.github.io/archivist) R package for archivisation of objects [has exemplified](http://pbiecek.github.io/archivist/justGetIT.html) its functionality by using eurostat to plot the number of people killed by road accidents, showing a decreasing trend of road accidents in many countries.

```{r 2015-manu-roadaccidents, echo=FALSE, eval=TRUE, fig.width=6, fig.height=5, fig.cap=fig$cap("2015-manu-roadaccidents", "Road accidents.")}
t1 <- get_eurostat("tsdtr420") %>%
  dplyr::filter(geo %in% c("UK", "SK", "FR", "PL", "ES", "PT", "LV"))

ggplot(t1, aes(x = time, y = values, color=geo, group=geo, shape=geo)) +
  geom_point(size=4) +
  geom_line() + theme_bw() + ggtitle("People killed in road accidents")
```


We can also look at the distribution of BMI between different age groups:

```{r 2015-manu-bmi, echo=FALSE, eval=TRUE, fig.width=6, fig.height=5} 
summary(tmp1 <- get_eurostat("hlth_ehis_de1", time_format = "raw"))
tmp1 %>%
  dplyr::filter( isced97 == "TOTAL" ,
          sex != "T",
          age != "TOTAL", geo == "PL") %>%
  ggplot(aes(y=values, x=age, fill=bmi)) +
  geom_bar(stat="identity") +
  facet_wrap(~sex) + coord_flip() +
  theme(legend.position="top")
```
 


## Discussion

The eurostat R package provides convenient tools to access open data
from Eurostat. When automated access to the data sets is integrated
with data analytical tools from other packages, this allows a seamless
automation of the data analytical process from raw data access to
statistical analysis and the final publication.

The package source code can be freely used, modified and distributed
under the BSD-2-clause (modified FreeBSD) license. A reproducible
version of this article is available at .. and can be used to generate
the manuscript text along with up-to-date figures and tables with the
latest version of the eurostat data. Manuscript automation provides
transparent documentation with full algorithmic details on how to
access, preprocess, analyse, and report data and analyses, thus
serving a template for good reproducible research practice. The reproducible source code for this manuscript is available at [eurostat github page](https://github.com/rOpenGov/eurostat/blob/master/vignettes/manuscript.Rmd).

The package exemplifies also the challenges and possible solutions to
reproducible research and automated open data retrieval. Possible
future extensions and improvements include design of specific data
representation class structures. This could facilitate harmonization
of the data representation with similar governmental data sets and
subsequent tool development. The latest development version of the
package can be installed from Github by following the instructions at
the [github site](https://github.com/rOpenGov/eurostat). We welcome
issues, bug reports and other feedback via [this development
site](https://github.com/ropengov/eurostat).


## Acknowledgements

We are grateful to [Eurostat](http://ec.europa.eu/eurostat/) for
maintaining the open data portal and the
[rOpenGov](https://github.ropengov.io) collection for supporting the
package development. This work has been partially funded by Academy of
Finland (decision ...). We also wish to thank Juuso Parkkinen and Joona Lehtomaki for their valuable feedback on this work.


## References

```{r, echo=FALSE, message=FALSE}
#You can embed citations, for example: `r citep(bib[["lahti14natcomm"]])`
#You can embed citations, for example2: @lahti14natcomm
#Cite with DOI: `r citep("10.1890/11-0011.1")`
#Cite URL `r citep("http://knowledgeblog.org/greycite")`
#For automated markdown citations, check [this](http://rmarkdown.rstudio.com/authoring_bibliographies_and_citations.html), [this](https://github.com/cboettig/knitcitations), and [this](http://www.carlboettiger.info/2012/03/24/citations-in-markdown-using-knitr.html). 
#write.bibtex(file="references.bib")
```

```{r, echo=FALSE, message=FALSE, results='asis'}
bibliography()
```


