---
title: "Vignette for the eurostat R package"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
    %\VignetteIndexEntry{Vignette for the eurostat R package}
    %\VignetteEngine{knitr::rmarkdown}
    %\VignetteDepends{Cairo}
    %\VignetteEncoding{UTF-8}
    \usepackage[utf8]{inputenc}
editor_options:
    chunk_output_type: console
---

# R Tools for Eurostat Open Data

This [rOpenGov](http://ropengov.github.io) R package provides tools to access [Eurostat database](https://ec.europa.eu/eurostat/data/database), which you can also browse on-line for the data sets and documentation. For contact information and source code, see the [package website](http://ropengov.github.io/eurostat/).

```{r setup, include=FALSE}
# Global options
library(knitr)
# opts_chunk$set(fig.path="fig/")
```


# Installation

Release version [(CRAN)](https://CRAN.R-project.org/package=eurostat):


```{r install, eval=FALSE}
install.packages("eurostat")
```

Development version [(Github)](https://github.com/rOpenGov/eurostat):

```{r install2, eval=FALSE}
library(devtools)
install_github("ropengov/eurostat")
```

```{r, echo=FALSE}
library(eurostat)
```

For examples on how to use the package, see the [online tutorial](http://ropengov.github.io/eurostat/) (Articles section of the homepage).

