---
title: "Examples on eurostat R package"
author: Leo Lahti, Janne Huovari, Markus Kainu, Przemyslaw Biecek
date: "`r Sys.Date()`"
bibliography: 
- references.bib
output: 
  md_document:
    variant: markdown_github
---
<!--
%\VignetteEngine{knitr::rmarkdown}
%\VignetteIndexEntry{eurostat Markdown Vignette}
%\usepackage[utf8]{inputenc}
-->

This document provides reproducible documentation to generate the
figures and tables for [our manuscript (in
preparation)](RJwrapper.pdf) introducing the [eurostat R
package](https://github.com/rOpenGov/eurostat). We assume below that
the required R extensions have already been installed.

To reproduce the manuscript figures and tables, clone the [eurostat
repository](https://github.com/rOpenGov/eurostat), navigate to the
[./vignettes/2015-RJournal](https://github.com/rOpenGov/eurostat/tree/master/vignettes/2015-RJournal)
subdirectory and convert the [Rmarkdown source
code](lahti-huovari-kainu-biecek.Rmd) in R with:

```{r 2015-manu-generate, echo=TRUE, message=FALSE, eval=FALSE}
library(knitr)
knit("lahti-huovari-kainu-biecek.Rmd")
```

This reproduces the manuscript figures as PNG images in the working
directory. To reproduce the complete manuscript PDF, navigate in the
[vignettes/2015-RJournal](https://github.com/rOpenGov/eurostat/blob/master/vignettes/2015-RJournal/)
folder, and run in R:

```{r 2015-manu-manuscript, echo=TRUE, message=FALSE, eval=FALSE}
source("main.R")
```


Let us first load some external R packages

```{r 2015-manu-init0, echo=TRUE, message=FALSE}
# Load libraries
library(eurostat)
library(knitr)
library(xtable)
library(tidyr)
library(plotrix)
library(ggplot2)

# Set ggplot theme
theme_set(theme_bw(20))

# Set figure folder
knitr::opts_chunk$set(fig.path = "./")
```

## Installation

Installing the CRAN release version:

```{r 2015-manu-install, echo=TRUE, message=FALSE, eval=FALSE}
install.packages("eurostat")
```

Installing the Github development version:

```{r 2015-manu-install-dev, echo=TRUE, message=FALSE, warning=FALSE, eval=FALSE}
library(devtools)
install_github("ropengov/eurostat")
```

## Search and download

To retrieve data for 'road accidents', for
instance, use:

```{r 2015-manu-search, echo=TRUE, message=FALSE, warning=FALSE}
library(eurostat)
query <- search_eurostat("road accidents", type = "table")
```

Investigate the first entry of our query:

```{r 2015-manu-search2, echo=TRUE, message=FALSE, warning=FALSE}
query$code[[1]]
query$title[[1]]
```


To retrieve the data set with this identifier, use:

```{r 2015-manu-get, echo=TRUE, message=FALSE}
dat <- get_eurostat(id = "tsdtr420", time_format = "num")
```

This produces a table:

```{r 2015-manu-getdata, echo=TRUE, message=FALSE}
#print(xtable(head(dat), label = "tab:getdatatable"))
kable(head(dat))
```


Same with human-readable labels:

```{r 2015-manu-getdata-labeled, echo=TRUE, message=FALSE}
# Convert into human readable labels
datl <- label_eurostat(dat)

# Print the table
#print(xtable(head(datl), label = "tab:getdatatable2"))
kable(head(datl))
```


## Road accidents visualization

The original detailed treatment of this example is provided in the related
[blog post](http://pbiecek.github.io/archivist/justGetIT.html).

```{r 2015-manu-roadacc, echo=TRUE, message=FALSE, warning=FALSE, fig.width=4, fig.height=4}
t1 <- get_eurostat("tsdtr420") %>%
  dplyr::filter(geo %in% c("UK", "SK", "FR", "PL", "ES", "PT", "LV"))
t1$Country <- t1$geo
ggplot(t1, aes(x = time, y = values, color=Country, group=Country, shape=Country)) +
  geom_point(size=4) + 
  geom_line() + theme_bw() + ggtitle("Road accidents")+
  xlab("Year") + ylab("Victims (n)") + theme(legend.position="top")
```

## Production of renewable energy

```{r 2015-manu-energy, echo=TRUE, message=FALSE, warning=FALSE, fig.width=8, fig.height=8}
dict <- c("Solid biofuels (excluding charcoal)" = "Biofuels",
          "Biogasoline" = "Biofuels",
          "Other liquid biofuels" = "Biofuels",
          "Biodiesels" = "Biofuels",
          "Biogas" = "Biofuels",
          "Hydro power" = "Hydro power",
          "Tide, Wave and Ocean" = "Hydro power",
          "Solar thermal" = "Wind, solar, waste and Other",
          "Geothermal Energy" = "Wind, solar, waste and Other",
          "Solar photovoltaic" = "Wind, solar, waste and Other",
          "Municipal waste (renewable)" = "Wind, solar, waste and Other",
          "Wind power" = "Wind, solar, waste and Other",
          "Bio jet kerosene" = "Wind, solar, waste and Other")

energy3 <- get_eurostat("ten00081") %>%
  label_eurostat(dat) %>% 
  filter(time == "2013-01-01",
         product != "Renewable energies") %>%
  mutate(nproduct = dict[as.character(product)], # just three categories
         geo = gsub(geo, pattern=" \\(.*", replacement="")) %>%
  select(nproduct, geo, values) %>% 
  group_by(nproduct, geo) %>%
  summarise(svalue = sum(values)) %>%
  group_by(geo) %>%
  mutate(tvalue = sum(svalue),
         svalue = svalue/sum(svalue)) %>%
  filter(tvalue > 1000,
         !grepl(geo, pattern="^Euro")) %>% # only large countrie
  spread(nproduct, svalue)

par(cex=1.5)
plotrix::triax.plot(as.matrix(energy3[, c(3,5,4)]),
                      show.grid = TRUE,
                      label.points = TRUE, point.labels = energy3$geo,cex.ticks=0.75,col.symbols = "red4",
                      pch = 19)
```


## Body-mass index

```{r 2015-manu-bmi, echo=TRUE, message=FALSE}
library(dplyr)
tmp1 <- get_eurostat("hlth_ehis_de1", time_format = "raw")
tmp1 %>%
  dplyr::filter( isced97 == "TOTAL" ,
          sex != "T",
          age != "TOTAL", geo == "PL") %>%
  mutate(BMI = factor(bmi, 
                      levels=c("LT18P5","18P5-25","25-30","GE30"), 
                      labels=c("<18.5", "18.5-25", "25-30",">30"))) %>%
  arrange(BMI) %>%
  ggplot(aes(y=values, x=age, fill=BMI)) +
  geom_bar(stat="identity") +
  facet_wrap(~sex) + coord_flip() +
  theme(legend.position="top") + ggtitle("Body mass index (BMI) by sex and age")+xlab("% of population")+scale_fill_brewer(type = "div")
```


## Map visualization

The source code for the detailed map visualization is hidden but [available](https://github.com/rOpenGov/eurostat/blob/master/vignettes/2015-RJournal/lahti-huovari-kainu-biecek.Rmd). For a detailed treatment of this example, see our [related blog post](http://ropengov.github.io/r/2015/05/01/eurostat-package-examples/).

```{r 2015-manu-mapexample, echo=FALSE, message=FALSE}
# Clean the cache
clean_eurostat_cache()

# Load dependencies
PACKAGES <- c("rgdal","maptools","rgeos","stringr","scales","grid")
tmp <- lapply(PACKAGES, require, character.only=TRUE)

# Downloading and manipulating the tabular data
df <- get_eurostat("tgs00026", time_format = "raw")

# convert time column from Date to numeric
df$time <- eurotime2num(df$time)

# subset time to have data for 
df <- df[df$time %in% c(2005,2010,2011),]

# spread the data into wide format
library(tidyr)
dw <- spread(df, time, values)

# Downloading and manipulating the spatial data
# We download the zipped shapefile in 1:60 million scale from year
# 2010 and subset it at the level of NUTS2.
# Load the GISCO shapefile
f <- download.file("http://ec.europa.eu/eurostat/cache/GISCO/geodatafiles/NUTS_2010_60M_SH.zip", destfile="NUTS_2010_60M_SH.zip")
# unzip
uz <- unzip("NUTS_2010_60M_SH.zip")

# read into SpatialPolygonsDataFrame
library(rgdal)
map <- readOGR(dsn = "./NUTS_2010_60M_SH/Data", layer = "NUTS_RG_60M_2010", verbose=FALSE)

# subset the spatialpolygondataframe at NUTS2-level
map_nuts2 <- subset(map, STAT_LEVL_ == 2)

# Joining tabular data with spatial data
# Third, we will make the both datas of same length,
# give then identical rownames and then merge the tabular data with the
# spatial data.

# Spatial dataframe has 467 rows and attribute data 275.
# We need to make attribute data to have similar number of rows
NUTS_ID <- as.character(map_nuts2$NUTS_ID)
VarX <- rep(NA, 316)
dat <- data.frame(NUTS_ID,VarX)

# then we shall merge this with Eurostat data.frame
dat2 <- merge(dat, dw, by.x="NUTS_ID", by.y="geo", all.x=TRUE)

# merge this manipulated attribute data with the spatialpolygondataframe
# rownames
row.names(dat2) <- dat2$NUTS_ID
row.names(map_nuts2) <- as.character(map_nuts2$NUTS_ID)

# order data
dat2 <- dat2[order(row.names(dat2)), ]
map_nuts2 <- map_nuts2[order(row.names(map_nuts2)), ]

# join
library(maptools)
dat2$NUTS_ID <- NULL
shape <- spCbind(map_nuts2, dat2)

# Preparing the data for ggplot2 visualization
## fortify spatialpolygondataframe into data.frame
library(ggplot2)
library(rgeos)
shape$id <- rownames(shape@data)
map.points <- fortify(shape, region = "id")
map.df <- merge(map.points, shape, by = "id")

# As we want to plot map faceted by years from 2003 to 2011
# we have to melt it into long format
# (variable with numerical names got X-prefix during the spCbind-merge,
# therefore the X-prefix in variable names)
library(tidyr)

# lets convert unit variable (that is a list) into character
map.df$unit <- as.character(map.df$unit)
map.df.l <- gather(map.df, "year", "value", 15:17)

# year variable (variable) is class string and type X20xx.
# Lets remove the X and convert it to numerical
library(stringr)
map.df.l$year <- str_replace_all(map.df.l$year, "X","")
map.df.l$year <- factor(map.df.l$year)
map.df.l$year <- as.numeric(levels(map.df.l$year))[map.df.l$year]

# Plotting the maps using ggplot2
library(ggplot2)
library(scales)
library(grid)

# Creating a custom function for creating the breaks and makeing them look neat
categories <- function(x, cat = 5) {
  
  library(stringr)
  levs <- as.data.frame(as.character(levels(cut_interval(x, cat))))
  names(levs) <- "orig"
  levs$mod <- str_replace_all(levs$orig, "\\[", "")
  levs$mod <- str_replace_all(levs$mod, "\\]", "")
  levs$mod <- str_replace_all(levs$mod, "\\(", "")
  levs$lower <- gsub(",.*$","", levs$mod)
  levs$upper <- gsub(".*,","", levs$mod)
  
  levs$lower <- factor(levs$lower)
  levs$lower <- round(as.numeric(levels(levs$lower))[levs$lower],0)
  
  levs$upper <- factor(levs$upper)
  levs$upper <- round(as.numeric(levels(levs$upper))[levs$upper],0)
  
  levs$labs <- paste(levs$lower,levs$upper, sep=" - ")
  
  labs <- as.character(c(levs$labs))
  y <- cut_interval(x, cat, right = FALSE, labels = labs)
  y <- as.character(y)
  y[is.na(y)] <- "No Data"
  y <- factor(y, levels=c("No Data",labs[1:cat]))
}

# years for for loop - now show just one year
years <- unique(map.df.l$year)[[1]]

# Loop over the three years
for (year in years) {
  
  # subset data
  plot_map <- map.df.l[map.df.l$year == year,]
  # set the breaks
  plot_map$value_cat <- categories(plot_map$value)
  
  p <- ggplot(data=plot_map, aes(long,lat,group=group))
  p <- p + geom_polygon(data = map.df.l, aes(long,lat),fill=NA,colour="white",size = 1)
  p <- p + geom_polygon(aes(fill = value_cat),colour="white",size=.2)
  p <- p + scale_fill_manual(values=c("Dim Grey","#d7191c","#fdae61","#ffffbf","#a6d96a","#1a9641")) 
  p <- p + coord_map(project="orthographic", xlim=c(-22,34), ylim=c(35,70))
  p <- p + labs(title = paste0("Disposable household incomes in  ",year))
  p <- p +  theme(legend.position = c(0.03,0.40), 
                          legend.justification=c(0,0),
                          legend.key.size=unit(6,'mm'),
                          legend.direction = "vertical",
                          legend.background=element_rect(colour=NA, fill=alpha("white", 2/3)),
                          legend.text=element_text(size=12), 
                          legend.title=element_text(size=12), 
                          title=element_text(size=16), 
                          panel.background = element_blank(), 
                          plot.background = element_blank(),
                          panel.grid.minor = element_line(colour = 'Grey80', size = .5, linetype = 'solid'),
                          panel.grid.major = element_line(colour = 'Grey80', size = .5, linetype = 'solid'),
                          axis.text = element_blank(), 
                          axis.title = element_blank(), 
                          axis.ticks = element_blank(), 
                          plot.margin = unit(c(-3,-1.5, -3, -1.5), "cm"))
  p <- p + guides(fill = guide_legend(title = "EUR per Year",
                                     title.position = "top", 
                                     title.hjust=0))
}
print(p)
```




## Country code tables

```{r 2015-manu-searchb, echo=TRUE, message=FALSE}
# Load EFTA country listing
data(efta_countries)

# Print the table
#print(xtable(efta_countries))
kable(efta_countries)
```

