---
title: "Tutorial for the eurostat R package"
date: "`r Sys.Date()`"
editor_options:
    chunk_output_type: inline
output: 
  html_document: 
    toc: yes
---

# R Tools for Eurostat Open Data

This [rOpenGov](http://ropengov.github.io) R package provides tools to access [Eurostat database](http://ec.europa.eu/eurostat/data/database), which you can also browse on-line for the data sets and documentation. For contact information and source code, see the [package website](http://ropengov.github.io/eurostat/).

```{r setup, include=FALSE}
# Global options
library(knitr)
opts_chunk$set(fig.path = "fig/")
options(citation.bibtex.max=999)
```

# Installation

Release version [(CRAN)](https://CRAN.R-project.org/package=eurostat):

```{r install, eval=FALSE}
install.packages("eurostat")
```

Development version [(Github)](https://github.com/rOpenGov/eurostat):

```{r install_devel, eval=FALSE}
library(remotes)
remotes::install_github("ropengov/eurostat")
```

Alternatively, development versions (more specifically: development versions in the master branch of eurostat GitHub repository) can be installed with the help of R-universe:

```{r install_runiverse, eval=FALSE}
# Enable this universe
options(repos = c(
  ropengov = "https://ropengov.r-universe.dev",
  CRAN = "https://cloud.r-project.org"
))

install.packages("eurostat")
```

The package is loaded with the library function.

```{r, echo=FALSE}
library(eurostat)
```

Overall, the eurostat package includes the following user-facing functions:

```{r, echo=FALSE,comment=NA}
cat(paste0(library(help = "eurostat")$info[[2]], collapse = "\n"))
```

```{r check_internet}
evaluate <- curl::has_internet()
```


# Finding data

First stop for most researchers would be to browse the [Eurostat Data Browser website](https://ec.europa.eu/eurostat/databrowser/explore/all/all_themes) or other thematically arranged sections in the Eurostat website. However, the eurostat R package offers some ways to search for datasets without leaving the R interface.

Function `get_eurostat_toc()` downloads a table of contents (TOC) of eurostat datasets.

```{r get_eurostat_toc, warning=FALSE, message=FALSE, eval = evaluate}
# Load the package
library(eurostat)

# Get Eurostat data listing
toc <- get_eurostat_toc()

# Check the first items
library(knitr)
kable(tail(toc))
```

The values in column 'code' are unique identifiers for each dataset that have to be used when downloading specific datasets. In the `get_eurostat()` function the dataset code is put into the first argument of the function, `id`.

From eurostat version 4.0.0 onwards the returned TOC object has had an additional column, `hierarchy`. It is used to determine which dataset belongs to which folder. This is helpful for example when downloading datasets in a single folder all at once.

From eurostat version 4.0.0 onwards it is possible to download TOC objects in French and German as well, in addition to English, which remains the default option. This enables new functionalities in other eurostat functions that have used the TOC object internally but retains backwards-compatibility with old code as the `lang` argument is not mandatory and queries without it continue to deliver the English version of the TOC object.

```{r get_eurostat_toc2, warning=FALSE, message=FALSE, eval = evaluate}
kable(head(get_eurostat_toc(lang = "fr")))
```

With `search_eurostat()` you can search the table of contents for particular patterns, e.g. all datasets related to *passenger transport*. With the `type` argument the user can choose which types of datasets the search should return: datasets, tables, folders or all types (the default).

According to [Eurostat database basic terminology](https://ec.europa.eu/eurostat/statistics-explained/index.php?title=Glossary:Eurostat_database_-_basic_terminology) "tables (predefined tables) are used to provide easy access to the main statistical indicators. They are based in general on datasets and are derived from them. They are predefined, non-modifiable and presented as two or three dimensional tables." The more general purpose datasets are, on the other hand, described to be "multi-dimensional tables" that have "up to 8 dimensions" and are used "store the base data, more appropriate for use by statistical and other experts via special applications".

```{r search_eurostat, warning=FALSE, message=FALSE, eval = evaluate}
# info about passengers
kable(head(search_eurostat("passenger transport")))
```

From eurostat version 4.0.0 onwards it possible to perform searches also from dataset codes. This is done by specifying the search column by setting the `column` attribute to `"code"`. Searching for codes can be useful in finding datasets that belong to the same folder or are part of a larger theme that shares similar dataset code patterns, such as "migr" for migration related statistics and "tran" in the case of (multimodal) transport statistics.

```{r search_eurostat2, warning=FALSE, message=FALSE, eval = evaluate}
kable(head(search_eurostat("migr", column = "code")))
```

Another new addition in version 4.0.0 is the option to perform searches from French and German language TOC versions as well by setting the `lang` argument to `"fr"` or `"de"`. Naturally, dataset codes are shared between language versions so French and German language searches should be conducted only on the title column.

```{r search_eurostat3, warning=FALSE, message=FALSE, eval = evaluate}
kable(head(search_eurostat("flughafen", column = "title", lang = "de")))
```

As mentioned in the beginning, codes for different dataset can be found also from the [Eurostat database](http://ec.europa.eu/eurostat/data/database). The Eurostat database gives codes in the Data Navigation Tree in parenthesis after the full name of the dataset, folder, or table.

# Downloading data 

The package supports two of the Eurostats download methods: the SDMX 2.1 REST API and the API Statistics ("JSON API"). The bulk download facility is the fastest method to download whole datasets. To download only a small section of the dataset the JSON API is faster, as it allows to make a data selection before downloading.

The end user does not usually have to bother where original data is downloaded, as both data sources are accessed via the main download function `get_eurostat()`. If only the table id is given, the whole table is downloaded from the SDMX 2.1 REST API. If any filters are given the JSON API is used instead. However, the `get_eurostat_json()` function used internally is also a user-facing function so that can be used as well.

We will use the dataset 'Modal split of air, sea and inland passenger transport' as an example dataset in this vignette. This is the percentage share of each mode of transport in total inland transport, expressed in passenger-kilometres (pkm) based on transport by passenger cars, buses and coaches, trains, aircraft, and seagoing vessels. All data should be based on movements on national territory, regardless of the nationality of the vehicle. However, the data collection is not harmonized at the EU level. For more detailed information about the dataset, see [Reference Metadata](https://ec.europa.eu/eurostat/cache/metadata/en/tran_hv_ms_esms.htm).

Pick and print the id of the data set to download: 
```{r get_id, warning=FALSE, message=FALSE, results='asis', eval = evaluate}
# Perform search, the output is a table of search results
search_results <- search_eurostat("Modal split of air, sea and inland passenger transport",
  type = "dataset"
)

# Since our search term was so detailed, we should have only 1 result / 1 row
kable(head(search_results))

# Get the id from the table
id <- search_results$code[1]

# Check the id
print(id)
```


```{r reevaluate, include = FALSE, eval = evaluate}
if (!is.na(id)) evaluate <- TRUE else evaluate <- FALSE
```

Get the whole corresponding table. As the table is annual data, it is more
convenient to use a numeric time variable than use the default date format, where yearly data would be coerced to be on the first day of the year (e.g. 2000-01-01).

```{r get_eurostat, warning=FALSE, message=FALSE, results='asis', eval = evaluate}
dat <- get_eurostat(id, time_format = "num", stringsAsFactors = TRUE)
```

The structure of the downloaded data set can be investigated by using the base R `str()` function:

```{r str_dat, warning=FALSE, message=FALSE, eval = evaluate}
str(dat)
```

```{r head_dat, warning=FALSE, message=FALSE, results='asis', eval = evaluate}
kable(head(dat))
```

You can get only a part of the dataset by defining `filters` argument. It
should be named list, where names corresponds to variable names (lower case) and
values are vectors of codes corresponding desired series (upper case). For
time variable, in addition to a `time` or `TIME_PERIOD` , also `sinceTimePeriod`, `untilTimePeriod` and a `lastTimePeriod` can be used. 

More information about filtering can be found in `get_eurostat()` and `get_eurostat_json()` function documentation.

```{r get_eurostat_json, warning=FALSE, message=FALSE, results='asis', eval=evaluate}
dat2 <- get_eurostat(id, filters = list(geo = c("EU27_2020", "FI"), lastTimePeriod = 1), time_format = "num")
kable(dat2)
```

## Replacing codes with labels

By default variables are returned as Eurostat codes, but to get human-readable 
labels instead, use a `type = "label"` argument in `get_eurostat()`.  

```{r json_labels, warning=FALSE, message=FALSE, results='asis', eval = evaluate}
dat_labeled2 <- get_eurostat(id,
  filters = list(
    geo = c("EU27_2020", "FI"),
    lastTimePeriod = 1
  ),
  type = "label", time_format = "num"
)
kable(head(dat_labeled2))
```

Eurostat codes in the downloaded data set can be replaced with
human-readable labels from the Eurostat dictionaries with the
`label_eurostat()` function.

```{r labels, warning=FALSE, message=FALSE, results='asis', eval = evaluate}
dat_labeled <- label_eurostat(dat)
kable(head(dat_labeled))
```

The `label_eurostat_vars()` allows conversion of variable names as well.

```{r name_labels, eval = evaluate}
print(label_eurostat_vars(id = "tran_hv_ms_psmod", names(dat_labeled)))
```


Vehicle information has 5 levels. You can check them now with:

```{r vehicle_levels, eval = evaluate}
levels(dat_labeled$vehicle)
```

## Downloading data interactively

New function in the eurostat package version 4.0.0 is the `get_eurostat_interactive()` function that allows users to search and download datasets with the help of interactive menus. If the user already knows which dataset they want to download, the `get_eurostat_interactive()` function can also take a dataset code as a parameter, skipping the search part of the interactive menu. Below we will demonstrate the whole process from search to download to printing a citation for the dataset, utilizing several different eurostat package functions at once.

```{text}
> get_eurostat_interactive()
Select language 

1: English
2: French
3: German

Selection: 1
Enter search term for data: aviation
Which dataset would you like to download?                                                             

1: [tran_sf_aviagah] Air accident victims in general aviation, by country of occurrence and country of registration of aircraft - maximum take-off mass above 2250 kg (source: EASA)
2: [tran_sf_aviagal] Air accident victims in general aviation by country of occurrence and country of registration of aircraft - maximum take-off mass under 2250 kg (source: EASA)
3: [avia_ec_enterp] Number of aviation and airport enterprises
4: [avia_ec_emp_ent] Employment in aviation and airport enterprises by sex


Selection: 4
Download the dataset? 

1: Yes
2: No

Selection: 1
Would you like to use default download arguments or set them manually? 

1: Default
2: Manually selected

Selection: 1
trying URL 'https://ec.europa.eu/eurostat/api/dissemination/sdmx/2.1/data/avia_ec_emp_ent?format=TSV&compressed=true'
Content type 'text/tab-separated-values; charset=UTF-8' length 1354 bytes
==================================================
downloaded 1354 bytes

Table avia_ec_emp_ent cached at /var/folders/f4/h_r3y60n0nn0qm6qx5hnx1s00000gn/T//RtmpDJ1gUA/eurostat/60ee371bcdcc9b130a20514d1e0d574d.rds
Print dataset citation? 

1: Yes
2: No

Selection: 1
Print code for downloading dataset? 

1: Yes
2: No

Selection: 1
Print dataset fixity checksum? 

1: Yes
2: No

Selection: 1
##### DATASET CITATION:

@Misc{avia-ec-emp-ent-2016-4-20,
  title = {Employment in aviation and airport enterprises by sex (avia\_ec\_emp\_ent)},
  url = {https://ec.europa.eu/eurostat/web/products-datasets/product?code=avia_ec_emp_ent},
  language = {english},
  year = {2016},
  author = {{Eurostat}},
  urldate = {2023-12-19},
  type = {Dataset},
  note = {Accessed 2023-12-19, dataset last updated 2016-04-20},
}

##### DOWNLOAD PARAMETERS:

get_eurostat(id = 'avia_ec_emp_ent')

##### FIXITY CHECKSUM:

Fixity checksum (md5) for dataset avia_ec_emp_ent: 36975282eaaea50a6e5f0e6cd64ef4d2

# A tibble: 450 × 6
   freq  enterpr sex   geo   TIME_PERIOD values
   <chr> <chr>   <chr> <chr> <date>       <dbl>
 1 A     AIRP    F     CY    2006-01-01     192
 2 A     AIRP    F     CY    2007-01-01     240
 3 A     AIRP    F     CY    2008-01-01     514
 4 A     AIRP    F     CY    2009-01-01    3278
 5 A     AIRP    F     CY    2010-01-01    2587
 6 A     AIRP    F     CY    2011-01-01    2255
 7 A     AIRP    F     CY    2012-01-01    2954
 8 A     AIRP    F     CZ    2001-01-01       0
 9 A     AIRP    F     CZ    2002-01-01       0
10 A     AIRP    F     CZ    2003-01-01       0
# ℹ 440 more rows
# ℹ Use `print(n = ...)` to see more rows
```

# Selecting and modifying data

## EFTA, Eurozone, EU and EU candidate countries

To facilitate smooth visualization of standard European geographic areas, the package provides ready-made lists of the country codes used in the eurostat database for EFTA (efta\_countries), Euro area (ea\_countries), EU (eu\_countries) and EU candidate countries (eu\_candidate\_countries). These can be used to select specific groups of countries for closer investigation. For conversions with other standard country coding systems, see the [countrycode](https://CRAN.R-project.org/package=countrycode) R package. To retrieve the country code list for EFTA, for instance, use:

```{r, echo=TRUE, eval = evaluate}
data(efta_countries)
kable(efta_countries)
```


## EU data from 2012 in all vehicles:

```{r eu_12, eval = evaluate}
dat_eu12 <- subset(dat_labeled, geo == "European Union - 27 countries (from 2020)" & TIME_PERIOD == 2012)
kable(dat_eu12, row.names = FALSE)
```

## EU data from 2008 - 2020 with vehicle types as variables:

Reshaping the data is best done with `spread()` in `tidyr`.
```{r eu_vehicles_table, eval = evaluate}
library("tidyr")
dat_eu_0012 <- subset(dat, geo == "EU27_2020" & TIME_PERIOD %in% c(2008:2020))
dat_eu_0012_wide <- spread(dat_eu_0012, vehicle, values)
kable(subset(dat_eu_0012_wide, select = -geo), row.names = FALSE)
```

## Train passengers for selected EU countries in 2008 - 2020

```{r trains_table, eval = evaluate}
dat_trains <- subset(dat_labeled, geo %in% c("Austria", "Belgium", "Finland", "Sweden") &
  TIME_PERIOD %in% c(2008:2020) &
  vehicle == "Trains")
dat_trains_wide <- spread(dat_trains, geo, values)
kable(subset(dat_trains_wide, select = -vehicle), row.names = FALSE)
```


## Other packages

### Strongly recommended

The `giscoR` ([package homepage](https://ropengov.github.io/giscoR/)) package used to be only suggested but starting from `eurostat` version 4.0.0 it has become a dependency of eurostat and required for using geospatial data functions. In addition to using `get_eurostat_geospatial()` from the `eurostat` package, it is highly recommended to study `giscoR` package functions and vignettes for creating more sophisticated visualisations to support geospatial analyses.

### Packages with similar functionalities

The [restatapi](https://CRAN.R-project.org/package=restatapi) R package has similar functionalities and some familiar function names for seasoned eurostat R package users. The `restatapi` package focuses more on statistical data and retrieving and returning data in a non-tidy data format.

The [rsdmx](https://CRAN.R-project.org/package=rsdmx) and [rjsdmx](https://CRAN.R-project.org/package=RJSDMX) R packages provide a more generic method to download data from a wide variety of statistical data providers that utilize the Statistical Data and Metadata eXchange ([SDMX](https://sdmx.org)) standards.

# Further examples

For further examples, see articles in the [package homepage](http://ropengov.github.io/eurostat/).

# Citations and related work

### Citing the data sources

Eurostat data: cite [Eurostat](http://ec.europa.eu/eurostat/).

Administrative boundaries: cite EuroGeographics


### Citing the eurostat R package

For main developers and contributors, see the [package homepage](http://ropengov.github.io/eurostat).

This work can be freely used, modified and distributed under the
BSD-2-clause (modified FreeBSD) license:

```{r citation, message=FALSE, eval=TRUE, echo=TRUE}
citation("eurostat")
```


### Contact

For contact information, see the [package homepage](http://ropengov.github.io/eurostat).


# Version info

This tutorial was created with

```{r sessioninfo, message=FALSE, warning=FALSE}
sessioninfo::session_info()
```
