---
title: "Map examples for the eurostat R package"
date: "`r Sys.Date()`"
output: 
  html_document: 
    toc: yes
editor_options:
  markdown:
    wrap: 80
    canonical: true
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  tidy = "styler",
  dev = "ragg_png",
  warning = FALSE,
  dpi = 300,
  out.width = "100%"
)
```

```{r setup}
library(eurostat)
```

# R Tools for Eurostat Open Data: maps

This [rOpenGov](http://ropengov.github.io) R package provides tools to access
[Eurostat database](http://ec.europa.eu/eurostat/data/database), which you can
also browse on-line for the data sets and documentation. For contact information
and source code, see the [package website](http://ropengov.github.io/eurostat/).

See the vignette of eurostat (`vignette(package = "eurostat")`) for installation
and basic use.

## Maps

> NOTE: we recommend to check also the `giscoR` package
> (<https://dieghernan.github.io/giscoR/>). This is another API package that
> provides R tools for Eurostat geographic data to support geospatial analysis
> and visualization.

### Disposable income of private households by NUTS 2 regions at 1:60mln resolution using tmap

The mapping examples below use [`tmap`](https://github.com/mtennekes/tmap)
package.

```{r maps1-1}
library(dplyr)
library(eurostat)
library(sf)
library(tmap)

# Download attribute data from Eurostat
sp_data <- eurostat::get_eurostat("tgs00026", time_format = "raw") %>%
  # subset to have only a single row per geo
  filter(TIME_PERIOD == 2016, nchar(geo) == 4) %>%
  # categorise
  mutate(income = cut_to_classes(values, n = 5))

# Download geospatial data from GISCO
geodata <- get_eurostat_geospatial(nuts_level = 2, year = 2016)

# merge with attribute data with geodata
map_data <- inner_join(geodata, sp_data, by = "geo")
```

Construct the map

```{r map1ex}
# Create and plot the map
map1 <- tm_shape(geodata,
  projection = "EPSG:3035",
  xlim = c(2400000, 7800000),
  ylim = c(1320000, 5650000)
) +
  tm_fill("lightgrey") +
  tm_shape(map_data) +
  tm_polygons("income",
    title = "Disposable household\nincomes in 2016",
    palette = "Oranges"
  )

print(map1)
```

Interactive maps can be generated as well

```{r maps1-inter, out.height="420px"}
# Interactive
tmap_mode("view")
map1
```

```{r maps1-2}
# Set the mode back to normal plotting
tmap_mode("plot")
print(map1)
```

### Disposable income of private households by NUTS 2 regions in Poland with labels at 1:1mln resolution using tmap

```{r maps2}
library(eurostat)
library(dplyr)
library(sf)

# Downloading and manipulating the tabular data
print("Let us focus on year 2016 and NUTS-3 level")
euro_sf2 <- get_eurostat("tgs00026",
  time_format = "raw",
  filter = list(time = "2016")
) %>%
  # Subset to NUTS-3 level
  dplyr::filter(grepl("PL", geo)) %>%
  # label the single geo column
  mutate(
    label = paste0(label_eurostat(.)[["geo"]], "\n", values, "€"),
    income = cut_to_classes(values)
  )

print("Download geospatial data from GISCO")
geodata <- get_eurostat_geospatial(
  resolution = "01", nuts_level = 2,
  year = 2016, country = "PL"
)

# Merge with attribute data with geodata
map_data <- inner_join(geodata, euro_sf2, by = "geo")

# plot map
library(tmap)

map2 <- tm_shape(geodata) +
  tm_fill("lightgrey") +
  tm_shape(map_data, is.master = TRUE) +
  tm_polygons("income",
    title = "Disposable household incomes in 2014",
    palette = "Oranges", border.col = "white"
  ) +
  tm_text("NUTS_NAME", just = "center") +
  tm_scale_bar() +
  tm_layout(legend.outside = TRUE)
map2
```

### Disposable income of private households by NUTS 2 regions at 1:10mln resolution using ggplot2

```{r maps4}
# Disposable income of private households by NUTS 2 regions at 1:1mln res
library(eurostat)
library(dplyr)
library(ggplot2)
data_eurostat <- get_eurostat("tgs00026", time_format = "raw") %>%
  filter(TIME_PERIOD == 2018, nchar(geo) == 4) %>%
  # classifying the values the variable
  dplyr::mutate(cat = cut_to_classes(values))

# Download geospatial data from GISCO
data_geo <- get_eurostat_geospatial(
  resolution = "01", nuts_level = "2",
  year = 2016
)

# merge with attribute data with geodata
data <- left_join(data_geo, data_eurostat, by = "geo")

ggplot(data) +
  # Base layer
  geom_sf(fill = "lightgrey", color = "lightgrey") +
  # Choropleth layer
  geom_sf(aes(fill = cat), color = "lightgrey", linewidth = 0.1, na.rm = TRUE) +
  scale_fill_brewer(palette = "Oranges", na.translate = FALSE) +
  guides(fill = guide_legend(reverse = TRUE, title = "euro")) +
  labs(
    title = "Disposable household income in 2018",
    caption = "© EuroGeographics for the administrative boundaries
                Map produced in R with data from Eurostat-package http://ropengov.github.io/eurostat"
  ) +
  theme_light() +
  coord_sf(
    xlim = c(2377294, 7453440),
    ylim = c(1313597, 5628510),
    crs = 3035
  )
```

# Citations and related work

### Citing the data sources

Eurostat data: cite [Eurostat](http://ec.europa.eu/eurostat/).

Administrative boundaries: cite EuroGeographics

### Citing the eurostat R package

For main developers and contributors, see the [package
homepage](http://ropengov.github.io/eurostat).

This work can be freely used, modified and distributed under the BSD-2-clause
(modified FreeBSD) license:

```{r citation, message=FALSE}
citation("eurostat")
```

### Contact

For contact information, see the [package
homepage](http://ropengov.github.io/eurostat).

# Version info

This tutorial was created with

```{r sessioninfo, message=FALSE}
sessioninfo::session_info()
```
