---
title: "Map examples for the eurostat R package"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
---

# R Tools for Eurostat Open Data

This [rOpenGov](http://ropengov.github.io) R package provides tools to access [Eurostat database](http://ec.europa.eu/eurostat/data/database), which you can also browse on-line for the data sets and documentation. For contact information and source code, see the [package website](http://ropengov.github.io/eurostat/).


See eurostat vignette for installation and basic use.

```{r, echo=FALSE}
library(eurostat)
```

## Maps 

###  Disposable income of private households by NUTS 2 regions at 1:60mln resolution using tmap

The mapping examples below use [`tmap`](https://github.com/mtennekes/tmap) package.


```{r maps1-1, eval=TRUE, fig.width=8, fig.height=8}
library(dplyr)
library(eurostat)
library(sf)
library(tmap)

# Download attribute data from Eurostat
sp_data <- eurostat::get_eurostat("tgs00026",
		                  time_format = "raw",
                                  stringsAsFactors = FALSE) %>% 
  # subset to have only a single row per geo
  dplyr::filter(time == 2010, nchar(geo) == 4) %>% 
  # categorise
  dplyr::mutate(income = cut_to_classes(values, n = 5))

# Download geospatial data from GISCO
geodata <- get_eurostat_geospatial(output_class = "sf",
                                   resolution = "60",
				   nuts_level = 2,
				   year = 2013)

# merge with attribute data with geodata
map_data <- inner_join(geodata, sp_data)
```


Construct the map

```{r map1ex, eval=TRUE, warning=FALSE}
map1 <- tmap::tm_shape(geodata) +
  tmap::tm_fill("lightgrey") +
  tmap::tm_shape(map_data) +
  tmap::tm_grid() +
  tmap::tm_polygons("income", title = "Disposable household\nincomes in 2010",  
                    palette = "Oranges")
print(map1)  
```

Interactive maps can be generated as well

```{r maps1-2, eval=FALSE, fig.width=8, fig.height=8}
# Interactive
tmap_mode("view")
map1

# Set the mode back to normal plotting
tmap_mode("plot")
print(map1)
```

### Disposable income of private households by NUTS 2 regions in Poland with labels at 1:1mln resolution using tmap

```{r maps2, fig.width=8, fig.height=8, warning=FALSE}
library(eurostat)
library(dplyr)
library(sf)
library(RColorBrewer)

# Downloading and manipulating the tabular data
print("Let us focus on year 2014 and NUTS-3 level")
euro_sf2 <- get_eurostat("tgs00026", time_format = "raw",
                         stringsAsFactors = FALSE,
			 filter = list(time = "2014")) %>% 
 
  # Subset to NUTS-3 level
  dplyr::filter(grepl("PL",geo)) %>% 
  # label the single geo column
  mutate(label = paste0(label_eurostat(.)[["geo"]], "\n", values, "€"),
         income = cut_to_classes(values))

print("Download geospatial data from GISCO")
geodata <- get_eurostat_geospatial(output_class = "sf", resolution = "60", nuts_level = 2, year = 2013)

# Merge with attribute data with geodata
map_data <- inner_join(geodata, euro_sf2)

# plot map
map2 <- tm_shape(geodata) +
  tm_fill("lightgrey") +
  tm_shape(map_data, is.master = TRUE) +
  tm_polygons("income", title = "Disposable household incomes in 2014",
              palette = "Oranges", border.col = "white") + 
  tm_text("NUTS_NAME", just = "center") + 
  tm_scale_bar()
map2
```

### Disposable income of private households by NUTS 2 regions at 1:10mln resolution using spplot

```{r maps3, fig.width=8, fig.height=8, dev='CairoPNG'}
library(sp)
library(eurostat)
library(dplyr)
library(RColorBrewer)
dat <- get_eurostat("tgs00026", time_format = "raw", stringsAsFactors = FALSE) %>% 
  # subsetting to year 2014 and NUTS-2 level
  dplyr::filter(time == 2014, nchar(geo) == 4) %>% 
  # classifying the values the variable
  dplyr::mutate(cat = cut_to_classes(values))

# Download geospatial data from GISCO
geodata <- get_eurostat_geospatial(output_class = "spdf", resolution = "10", nuts_level = 2, year = 2013)

# merge with attribute data with geodata
geodata@data <- left_join(geodata@data, dat)

# plot map
sp::spplot(obj = geodata, "cat", main = "Disposable household income",
	   xlim = c(-22,34), ylim = c(35,70), 
           col.regions = c("dim grey", brewer.pal(n = 5, name = "Oranges")),
	   col = "white", usePolypath = FALSE)
```

### Disposable income of private households by NUTS 2 regions at 1:60mln resolution using ggplot2

Meanwhile the CRAN version of `ggplot2` is lacking support for simple features, you can plot maps with `ggplot2` by downloading geospatial data as `data.frame` with `output_class` argument set as `df`.

```{r maps4, fig.width=8, fig.height=8, dev='CairoPNG'}
library(eurostat)
library(dplyr)
library(ggplot2)
dat <- get_eurostat("tgs00026", time_format = "raw", stringsAsFactors = FALSE) %>% 
  # subsetting to year 2014 and NUTS-2 level
  dplyr::filter(time == 2014, nchar(geo) == 4) %>% 
  # classifying the values the variable
  dplyr::mutate(cat = cut_to_classes(values))

# Download geospatial data from GISCO
geodata <- get_eurostat_geospatial(resolution = "60", nuts_level = "2", year = 2013)

# merge with attribute data with geodata
map_data <- inner_join(geodata, dat)

ggplot(data=map_data) + geom_sf(aes(fill=cat),color="dim grey", size=.1) + 
    scale_fill_brewer(palette = "Oranges") +
  guides(fill = guide_legend(reverse=T, title = "euro")) +
  labs(title="Disposable household income in 2014",
       caption="(C) EuroGeographics for the administrative boundaries 
                Map produced in R with a help from Eurostat-package <github.com/ropengov/eurostat/>") +
  theme_light() + theme(legend.position=c(.8,.8)) +
  coord_sf(xlim=c(-12,44), ylim=c(35,70))

```


# Citations and related work

### Citing the data sources

Eurostat data: cite [Eurostat](http://ec.europa.eu/eurostat/).

Administrative boundaries: cite EuroGeographics


### Citing the eurostat R package

For main developers and contributors, see the [package homepage](http://ropengov.github.io/eurostat).

This work can be freely used, modified and distributed under the
BSD-2-clause (modified FreeBSD) license:

```{r citation, message=FALSE, eval=TRUE, echo=TRUE}
citation("eurostat")
```


### Contact

For contact information, see the [package homepage](http://ropengov.github.io/eurostat).


# Version info

This tutorial was created with

```{r sessioninfo, message=FALSE, warning=FALSE}
sessionInfo()
```
